---
title: Automated data-integrity checks
subtitle: SCR data checker app
author: |
  | Snow leopard SECR workflow workshop
  | Ian Durbach and Cornelia Oedekoven
institute: |
  | Centre for Research into Ecological & Environmental Modelling
  | University of St Andrews
fontsize: 10pt
format:
  beamer:
    theme: metropolis
    incremental: false
    slide-level: 2
    pdf-engine: xelatex
    syntax-highlighting: tango
classoption: t
urlcolor: teal
execute:
  echo: true
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(knitr)

knitr::opts_knit$set(global.par = TRUE)
knitr::opts_chunk$set(fig.path = "code-and-figures/")

def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(
  chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    chunk_size <- if (is.null(options$size)) "normalsize" else options$size
    ifelse(
      chunk_size != "normalsize",
      paste0("\n \\", chunk_size, "\n\n", x, "\n\n \\normalsize"),
      x
    )
  }
)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines) == 1) {
    if (length(x) > lines) {
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})
```

## Session objectives

\alert{Observation:} unstandardized, undocumented data entry practices introduce unnecessary errors and take a lot of time to correct.

\alert{Main goal:} check common data problems before building `traps` and `capthist` objects for `secr`.

1. Show where the app fits in the workflow
2. Summarize the main checks on `traps` and `detections` csv files
3. Show how to launch the app and use the report
4. Leave the details to the live demo

## Where this app fits

\alert{Key idea:} this is an initial check **before** the main SECR scripts.

- Upload a traps csv and a detections csv
- Let the app propose standard column names
- Approve the mapping if it looks right, else manually revise
- Review errors/warnings, correct where needed
- Download an HTML report for record-keeping

## What the app expects

The app is built around the standard workshop inputs:

- `traps.csv`: detector locations, effort, and optionally dates/covariates
- `detections.csv`: animal detections by session, occasion, and trap

It first checks column names and can map common alternatives such as:

- `trap id`, `camera id`, `station` $\rightarrow$ `trapID`
- `easting`, `northing` $\rightarrow$ `x`, `y`
- `latitude`, `longitude` $\rightarrow$ `lat`, `long`
- `animal id`, `individual id` $\rightarrow$ `animalID`

If no `session` column is found, the app can create `session = 1`.

## Checks on the traps file

Tasks involved:

- validate identifiers
- validate coordinates
- validate effort
- compare dates against effort
- summarize extra covariates

## Checks on the traps file

Main checks:

- `session` must be a positive integer, or text with no spaces
- `trapID` must be unique within session
- `trapID` with spaces or punctuation is flagged as a warning
- coordinates must be numeric, with no comma decimals or stray whitespace
- at least one coordinate pair must exist: `x/y` or `lat/long`
- duplicated coordinates used by different trap IDs are flagged
- possible swapped coordinates are flagged
- `effort` must be numeric and strictly positive
- `start_date` and `end_date` are parsed and compared with `effort`

## Checks on the detections file

Tasks involved:

- validate IDs and occasions
- parse dates
- compare detections against traps
- review optional covariates

## Checks on the detections file

Main checks:

- `session` must be a positive integer, or text with no spaces
- `animalID` must not contain spaces or punctuation
- placeholder IDs such as `unknown`, `unk`, `unsure`, `na` are flagged
- alternate forms of the same ID, such as notes in brackets, are flagged
- `occasion` must be a positive integer
- `date` is parsed and impossible dates are flagged
- overlapping date ranges across occasions are flagged
- optional `sex` values outside `Male/Female` or `M/F` are flagged

## Cross-file checks

Main checks:

- detections whose `trapID`/`session` combination is missing from `traps.csv`
- detections at traps with `effort <= 0`
- detection sessions missing from `traps.csv`
- detection dates outside the recorded trap operating period
- sessions present in `traps.csv` but absent from detections, shown in the summaries

Also gives quick summaries, plots, and maps to help spot problems visually.

## Installation and launch

Use the launcher script (recommended):

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
source("data_checker/run_data_checker.R")
```
\normalsize

Or from the `course_materials` R project:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
install.packages(c(
  "shiny", "shinydashboard", "DT", "ggplot2", "dplyr", "readr",
  "stringr", "tidyr", "scales", "leaflet", "sf", "rmarkdown", "htmltools"
))

shiny::runApp("data_checker")
```
\normalsize

## How to use it

Brief workflow for the demo:

1. Upload `traps.csv` and `detections.csv`, or pick one of the example files
2. Check the proposed column-name mapping and approve it
3. (Optional) Enter CRS `+proj=utm +zone=44 +datum=WGS84 +units=m +no_defs`
for map display
4. Open the Data Checks tab and work through errors before warnings
5. Download the HTML report once the inputs look sensible

\vspace{0.6cm}

\alert{Key idea:} fix input problems here, before they turn into harder-to-interpret errors later in `secr`.

## Live demo

- Upload a traps csv and a detections csv with required column names 
- If no exact match, app will propose which columns to use (but no renaming, yet)

\begin{center}
\includegraphics[width=0.75\textwidth]{data_checker_mapping_raw.png}
\end{center}

## Live demo

- Data Checks tab shows the main warnings and errors from both files.

\begin{center}
\includegraphics[width=0.75\textwidth]{data_checker_checks_raw.png}
\end{center}

## Live demo

- Bad example files are provided to demonstrate the kinds of problems the app catches.

\begin{center}
\includegraphics[width=0.75\textwidth]{data_checker_checks_bad.png}
\end{center}
