---
title: "Fit advanced SECR models"
author: "Ian Durbach & Cornelia Oedekoven"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: false
  pdf: default
format-links:
  - pdf
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  fig.align = "center",
  dpi = 150
)

input_file <- knitr::current_input()
if (!is.null(input_file)) {
  repo_root <- normalizePath(file.path(dirname(input_file), "../..") )
  knitr::opts_knit$set(root.dir = repo_root)
}
```

# Covered in this tutorial

- Fit full-survey advanced models
- Fit shortened-survey models using the same fitting mask
- Fit multisession models with session-specific masks and `userdist`
- Fit sex-specific full-survey models using `hcov = "sex"` and `groups = "sex"`
- Compare each candidate set with AIC
- Save fitted models for the results tutorial

# Introduction

This tutorial fits four related advanced model sets. The full-survey and shortened-survey analyses are closest to the basic model-fitting workflow, but they use the elevation-constrained fitting mask and gorge-avoiding movement distances. The multisession analysis uses a list of capture histories, masks, and `userdist` objects. The sex-specific analysis demonstrates two ways to include individual sex information.

The model sets are deliberately small and explicit. This keeps the teaching focus on how the advanced inputs change the fitting call, rather than on a large model-selection exercise.

# Preparations

Only `secr` is needed for fitting because the capture histories, masks, and user-distance objects have already been created.

```{r packages}
# Packages needed
library(secr)
```

# Encounter and detection functions

The advanced models use the hazard half-normal detection function, `detectfn = "HHN"`. With count detectors, it is often more useful to think about the encounter-rate function, $\lambda(d)$, because an individual can be detected more than once during an occasion. Detection probability, $g(d)$, is then derived from the encounter rate.

For the hazard half-normal model, encounter rate declines with distance from an activity centre to a camera. `lambda0` is the encounter-rate parameter at distance zero, and `sigma` controls the spatial scale over which encounter rate declines. Both are fitted on a link-function scale, so we usually interpret predictions on the natural scale rather than reading raw coefficients directly.

Trap covariates such as `cliff` and `valley_stream` enter the `lambda0` model. Mask covariates such as `std_tri` and `std_elev` enter the density model, `D`. Keeping those roles distinct is especially important in the advanced tutorial because the mask and detector covariates come from different spatial objects.

# Full-survey models

The full-survey model set is the closest advanced equivalent to the basic model-fitting tutorial. The two important changes are that the fitting mask is `mask_model`, which has been clipped to plausible elevations, and that `details = list(userdist = userd_model)` tells `secr.fit()` to use gorge-avoiding movement distances.

```{r full-load-inputs}
output_file <- "namkha_advanced/output/namkha_advanced_fitted_models_full.RData"

load("namkha_advanced/output/namkha_advanced_secr_inputs_capthist.RData")
load("namkha_advanced/output/namkha_advanced_secr_inputs_mask.RData")

# Check the capture history and covariates before fitting models.
summary(full_ch)
names(covariates(full_traps))
names(covariates(mask_model))
```

The fitted models use the hazard half-normal detection function. `D` is modelled with mask covariates such as ruggedness and elevation; `lambda0` is modelled with detector covariates such as cliff and valley/stream.

```{r fit-full-models}
f0 <- secr.fit(
  full_ch,
  detectfn = "HHN",
  mask = mask_model,
  model = list(D ~ 1, lambda0 ~ 1, sigma ~ 1),
  details = list(userdist = userd_model),
  verify = FALSE,
  trace = FALSE
)

f1 <- secr.fit(
  full_ch,
  detectfn = "HHN",
  mask = mask_model,
  model = list(D ~ std_tri, lambda0 ~ 1, sigma ~ 1),
  details = list(userdist = userd_model),
  verify = FALSE,
  start = f0,
  trace = FALSE
)

f2 <- secr.fit(
  full_ch,
  detectfn = "HHN",
  mask = mask_model,
  model = list(D ~ std_elev, lambda0 ~ 1, sigma ~ 1),
  details = list(userdist = userd_model),
  verify = FALSE,
  start = f0,
  trace = FALSE
)

f3 <- secr.fit(
  full_ch,
  detectfn = "HHN",
  mask = mask_model,
  model = list(D ~ 1, lambda0 ~ valley_stream, sigma ~ 1),
  details = list(userdist = userd_model),
  verify = FALSE,
  start = f0,
  trace = FALSE
)

f4 <- secr.fit(
  full_ch,
  detectfn = "HHN",
  mask = mask_model,
  model = list(D ~ std_tri, lambda0 ~ cliff + valley_stream, sigma ~ 1),
  details = list(userdist = userd_model),
  verify = FALSE,
  start = f1,
  trace = FALSE
)

AIC(f0, f1, f2, f3, f4)
```

When fitting many models, it is good practice to check whether the optimiser finished cleanly and whether any coefficient estimates are `NA` or infinite. `trace = FALSE` suppresses the iteration output from `secr.fit()`, but it does not remove the need to inspect warnings and fitted objects.

```{r check-full-model-fits}
full_fit_checks <- data.frame(
  model = c("f0", "f1", "f2", "f3", "f4"),
  fit_code = sapply(list(f0, f1, f2, f3, f4), function(x) x$fit$code),
  any_na_coef = sapply(list(f0, f1, f2, f3, f4), function(x) any(is.na(coefficients(x)))),
  any_inf_coef = sapply(list(f0, f1, f2, f3, f4), function(x) any(is.infinite(abs(unlist(coefficients(x))))))
)
full_fit_checks
```

The fitted coefficients returned by `coef()` are on the link scale. For positive parameters such as `lambda0` and `sigma`, the natural-scale estimates are obtained by applying the inverse link function. `predict()` does this transformation for the fitted model and is usually the safer way to inspect interpretable parameter estimates.

```{r inspect-full-null-coefficients}
coef(f0)
predict(f0)
```

The fitted models are saved separately from the shortened, multisession, and sex-specific model sets.

```{r save-full-models}
save(
  f0, f1, f2, f3, f4,
  file = output_file
)
```

# Shortened-survey models

The shortened-survey analysis is one response to closure concerns. It uses `short_ch` and `short_traps`, which already contain detections and effort restricted to the shorter survey window. The mask and `userdist` objects are the same as in the full-survey analysis so that the two analyses differ mainly in the data window.

```{r short-load-inputs}
output_file <- "namkha_advanced/output/namkha_advanced_fitted_models_short.RData"

load("namkha_advanced/output/namkha_advanced_secr_inputs_capthist.RData")
load("namkha_advanced/output/namkha_advanced_secr_inputs_mask.RData")

# Check the shortened capture history and covariates before fitting models.
summary(short_ch)
names(covariates(short_traps))
names(covariates(mask_model))
```

The candidate set mirrors the full-survey model set closely. This makes later comparisons easier to interpret.

```{r fit-short-models}
# This analysis uses the shorter survey period, while still retaining the
# elevation-constrained mask and the gorge-avoiding movement distances.

# Null model: density, encounter rate, and movement scale are constant.
s0 <- secr.fit(
  short_ch,
  detectfn = "HHN",
  mask = mask_model,
  model = list(D ~ 1, lambda0 ~ 1, sigma ~ 1),
  details = list(userdist = userd_model),
  verify = FALSE,
  trace = FALSE
)

# Density varies with terrain ruggedness.
s1 <- secr.fit(
  short_ch,
  detectfn = "HHN",
  mask = mask_model,
  model = list(D ~ std_tri, lambda0 ~ 1, sigma ~ 1),
  details = list(userdist = userd_model),
  verify = FALSE,
  start = s0,
  trace = FALSE
)

# Density varies with elevation.
s2 <- secr.fit(
  short_ch,
  detectfn = "HHN",
  mask = mask_model,
  model = list(D ~ std_elev, lambda0 ~ 1, sigma ~ 1),
  details = list(userdist = userd_model),
  verify = FALSE,
  start = s0,
  trace = FALSE
)

# Encounter rate varies by whether the camera is at a cliff.
s3 <- secr.fit(
  short_ch,
  detectfn = "HHN",
  mask = mask_model,
  model = list(D ~ 1, lambda0 ~ cliff, sigma ~ 1),
  details = list(userdist = userd_model),
  verify = FALSE,
  start = s0,
  trace = FALSE
)

# Combined model with ruggedness on density and local topography on encounter rate.
s4 <- secr.fit(
  short_ch,
  detectfn = "HHN",
  mask = mask_model,
  model = list(D ~ std_tri, lambda0 ~ cliff + valley_stream, sigma ~ 1),
  details = list(userdist = userd_model),
  verify = FALSE,
  start = s1,
  trace = FALSE
)

AIC(s0, s1, s2, s3, s4)
```

```{r save-short-models}
save(
  s0, s1, s2, s3, s4,
  file = output_file
)
```

# Multisession models

The multisession analysis uses `ch_ms`, `mask_ms`, and `userd_ms`. Each of these is a list aligned by session. The model parameters are shared across sessions unless a session term is included in the formula.

```{r ms-load-inputs}
output_file <- "namkha_advanced/output/namkha_advanced_fitted_models_ms.RData"

load("namkha_advanced/output/namkha_advanced_secr_inputs_capthist.RData")
load("namkha_advanced/output/namkha_advanced_secr_inputs_msmask.RData")

# Check the multisession capture history and covariates before fitting models.
summary(ch_ms)
names(covariates(traps_ms[[1]]))
names(covariates(mask_ms[[1]]))
```

The first five models use the same broad structure as the full-survey models. The sixth model, `ms5`, includes `D ~ session` as an illustration of session-specific density, but this creates complications for extrapolation and is excluded from the results workflow.

```{r fit-ms-models}
ms0 <- secr.fit(
  ch_ms,
  detectfn = "HHN",
  mask = mask_ms,
  model = list(D ~ 1, lambda0 ~ 1, sigma ~ 1),
  details = list(userdist = userd_ms),
  verify = FALSE,
  trace = FALSE
)

ms1 <- secr.fit(
  ch_ms,
  detectfn = "HHN",
  mask = mask_ms,
  model = list(D ~ std_tri, lambda0 ~ 1, sigma ~ 1),
  details = list(userdist = userd_ms),
  verify = FALSE,
  start = ms0,
  trace = FALSE
)

ms2 <- secr.fit(
  ch_ms,
  detectfn = "HHN",
  mask = mask_ms,
  model = list(D ~ std_elev, lambda0 ~ 1, sigma ~ 1),
  details = list(userdist = userd_ms),
  verify = FALSE,
  start = ms0,
  trace = FALSE
)

ms3 <- secr.fit(
  ch_ms,
  detectfn = "HHN",
  mask = mask_ms,
  model = list(D ~ 1, lambda0 ~ valley_stream, sigma ~ 1),
  details = list(userdist = userd_ms),
  verify = FALSE,
  start = ms0,
  trace = FALSE
)

ms4 <- secr.fit(
  ch_ms,
  detectfn = "HHN",
  mask = mask_ms,
  model = list(D ~ std_tri, lambda0 ~ cliff + valley_stream, sigma ~ 1),
  details = list(userdist = userd_ms),
  verify = FALSE,
  start = ms0,
  trace = FALSE
)

# Density varies by session, allowing different abundance in each time block.
# Note this model can't easily be used for extrapolation.
ms5 <- secr.fit(
  ch_ms,
  detectfn = "HHN",
  mask = mask_ms,
  model = list(D ~ session, lambda0 ~ 1, sigma ~ 1),
  details = list(userdist = userd_ms),
  verify = FALSE,
  start = ms0,
  trace = FALSE
)

AIC(ms0, ms1, ms2, ms3, ms4, ms5)
```

```{r save-ms-models}
save(
  ms0, ms1, ms2, ms3, ms4, ms5,
  file = output_file
)
```

# Sex-specific full-survey models

The final model-fitting script demonstrates two ways to include sex. Individual-covariate models use `hcov = "sex"` and can retain animals with unknown sex. Grouped models use `groups = "sex"` and require removing animals whose sex is unknown.

These models are saved to a separate file so they do not overwrite the main full-survey models.

```{r sex-load-inputs}
output_file <- "namkha_advanced/output/namkha_advanced_fitted_models_full_sex.RData"

load("namkha_advanced/output/namkha_advanced_secr_inputs_capthist.RData")
load("namkha_advanced/output/namkha_advanced_secr_inputs_mask.RData")

# Check the capture history and covariates before fitting sex-specific models.
summary(full_ch)
names(covariates(full_traps))
names(covariates(mask_model))

# h2 models only use first two levels of the sex covariate
levels(covariates(full_ch)$sex)
```

In `hcov` models, `h2` represents the first two levels of the individual covariate supplied through `hcov = "sex"`. Here it is used on `lambda0` so that encounter rate can differ by sex while density remains a population-level surface.

```{r fit-hcov-sex-models}
# The fitting mask is the trap-buffer mask after clipping to plausible elevation.
# Custom user distances are used so movement can avoid the gorge barrier.
# ADVANCED ISSUE: hcov = "sex" allows detection parameters to vary by individual
# sex while still retaining animals with unknown sex.

# Null model with sex as an individual covariate on detection.
f0 <- secr.fit(full_ch, detectfn = "HHN", mask = mask_model,
  model = list(D ~ 1, lambda0 ~ h2, sigma ~ 1),
  hcov = "sex", details = list(userdist = userd_model), verify = FALSE, trace = FALSE)

# Density varies with terrain ruggedness.
f1 <- secr.fit(full_ch, detectfn = "HHN", mask = mask_model,
  model = list(D ~ std_tri, lambda0 ~ h2, sigma ~ 1),
  hcov = "sex", details = list(userdist = userd_model), verify = FALSE, start = f0, trace = FALSE)

# Density varies with elevation.
f2 <- secr.fit(full_ch, detectfn = "HHN", mask = mask_model,
  model = list(D ~ std_elev, lambda0 ~ h2, sigma ~ 1),
  hcov = "sex", details = list(userdist = userd_model), verify = FALSE, start = f0, trace = FALSE)

# Encounter rate varies by whether the camera is in a valley or stream setting.
f3 <- secr.fit(full_ch, detectfn = "HHN", mask = mask_model,
  model = list(D ~ 1, lambda0 ~ valley_stream + h2, sigma ~ 1),
  hcov = "sex", details = list(userdist = userd_model), verify = FALSE, start = f0, trace = FALSE)

# Combined model with ruggedness on density and local topography on encounter rate.
f4 <- secr.fit(full_ch, detectfn = "HHN", mask = mask_model,
  model = list(D ~ std_tri, lambda0 ~ cliff + valley_stream + h2, sigma ~ 1),
  hcov = "sex", details = list(userdist = userd_model), verify = FALSE, start = f1, trace = FALSE)

AIC(f0, f1, f2, f3, f4)
```

Grouped models estimate separate real parameters by sex. Because unknown-sex animals are removed, these models should not be compared in the same AIC table as the `hcov` models.

```{r fit-grouped-sex-models}
# ADVANCED ISSUE: groups = "sex" estimates separate real parameters by sex, but
# it requires removing individuals whose sex is unknown.
full_ch_known_sex <- subset(full_ch, subset = covariates(full_ch)$sex != "Unknown")
full_ch_known_sex <- shareFactorLevels(full_ch_known_sex)

summary(full_ch, terse = TRUE)
summary(full_ch_known_sex, terse = TRUE)

# Null grouped model for known-sex individuals.
f0g <- secr.fit(full_ch_known_sex, detectfn = "HHN", mask = mask_model,
  model = list(D ~ 1, lambda0 ~ g, sigma ~ 1),
  groups = "sex", details = list(userdist = userd_model), verify = FALSE, trace = FALSE)

# Grouped model with density varying with terrain ruggedness.
f1g <- secr.fit(full_ch_known_sex, detectfn = "HHN", mask = mask_model,
  model = list(D ~ std_tri, lambda0 ~ g, sigma ~ 1),
  groups = "sex", details = list(userdist = userd_model), verify = FALSE, start = f0g, trace = FALSE)

# Grouped model with density varying with elevation.
f2g <- secr.fit(full_ch_known_sex, detectfn = "HHN", mask = mask_model,
  model = list(D ~ std_elev, lambda0 ~ g, sigma ~ 1),
  groups = "sex", details = list(userdist = userd_model), verify = FALSE, start = f0g, trace = FALSE)

# Grouped model with encounter rate varying at valley/stream cameras.
f3g <- secr.fit(full_ch_known_sex, detectfn = "HHN", mask = mask_model,
  model = list(D ~ 1, lambda0 ~ valley_stream + g, sigma ~ 1),
  groups = "sex", details = list(userdist = userd_model), verify = FALSE, start = f0g, trace = FALSE)

# Grouped combined model with density and encounter-rate covariates.
f4g <- secr.fit(full_ch_known_sex, detectfn = "HHN", mask = mask_model,
  model = list(D ~ std_tri, lambda0 ~ cliff + valley_stream + g, sigma ~ 1),
  groups = "sex", details = list(userdist = userd_model), verify = FALSE, start = f0g, trace = FALSE)

# Do not compare hcov and grouped models in the same AIC table because they use
# different data or parameterisations.
AIC(f0, f1, f2, f3, f4)
AIC(f0g, f1g, f2g, f3g, f4g)
```

```{r save-sex-models}
save(
  f0, f1, f2, f3, f4,
  f0g, f1g, f2g, f3g, f4g,
  file = output_file
)
```

# What comes next

The next tutorial reports the full-survey, shortened-survey, and multisession results. It estimates abundance and density over the elevation-constrained Namkha survey area, makes predicted density maps, and plots density and detection covariate effects.
