---
title: "Report advanced model results"
author: "Ian Durbach & Cornelia Oedekoven"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: false
  pdf: default
format-links:
  - pdf
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  fig.align = "center",
  dpi = 150
)

input_file <- knitr::current_input()
if (!is.null(input_file)) {
  repo_root <- normalizePath(file.path(dirname(input_file), "../..") )
  knitr::opts_knit$set(root.dir = repo_root)
}
```

# Covered in this tutorial

- Summarise AIC, abundance, density, and precision for fitted advanced models
- Report abundance over the elevation-constrained Namkha survey area
- Replace model `userdist` objects when reporting over a new mask
- Make predicted density maps for full, shortened, and multisession analyses
- Plot density covariate effects for the selected example models
- Plot detection functions for baseline, cliff, and valley/stream detector settings

# Introduction

This tutorial turns fitted advanced models into results tables, maps, and covariate-effect plots. The reporting workflow is the same in spirit as the basic tutorial, but the advanced models introduce one important extra step: because the fitted models used custom gorge-avoiding movement distances, reporting over a new mask requires a matching `userdist` object for that mask.

The full-survey, shortened-survey, and multisession workflows are similar. The differences are the fitted model file, the traps object used for plotting, and the way multisession abundance and density predictions are returned.

# Preparations

All three results workflows use the same plotting packages and the `plot_detfunction()` helper from the basic tutorial.

```{r packages}
# Packages needed
library(dplyr)
library(secr)
library(sf)
library(ggplot2)
library(MASS)

source("namkha_basic/code/more-utilities.R")
```

# Common reporting pattern

Model fitting used `mask_model`, the elevation-constrained trap-buffer mask. Reporting uses `mask_survey_area_elev`, the elevation-constrained Namkha survey-area mask. This keeps abundance and density on the biological and management region of interest, while still allowing the likelihood to account for activity centres outside the boundary.

For `region.N()`, the fitted model needs a `userdist` object whose columns match the reporting mask. The scripts therefore create `fitted_models_region`, a copy of the fitted model list with the user-distance object replaced by `userd_survey_area_elev`. The original fitted models are not refitted.

`region.N()` can return expected abundance, `E.N`, and realised abundance, `R.N`. Expected abundance is the integral of the fitted density surface over the region. Realised abundance is the number of individuals in a particular realisation of the spatial point process. For these reporting tables we use `E.N`, because it is the smoother model-based abundance estimate most commonly reported with density.

For maps, the mask starts as points because that is what `secr` uses for numerical integration. To display a density surface, the scripts convert each mask point to a square polygon centred on the point. The square extends half the mask spacing in each direction, so adjacent squares tile the mask region.

# Full-survey results

The full-survey results use the fitted models from `04a_model_fitting_full.R`. The reporting mask is the elevation-constrained Namkha survey area, and the matching `userdist` object is `userd_survey_area_elev`.

```{r full-load-inputs}
results_dir <- "namkha_advanced/output/results/full"

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)

load("namkha_advanced/output/namkha_advanced_secr_inputs_capthist.RData")
load("namkha_advanced/output/namkha_advanced_secr_inputs_mask.RData")
load("namkha_advanced/output/namkha_advanced_fitted_models_full.RData")

# Put fitted models in one named list for easy looping
fitted_models <- list(f0 = f0, f1 = f1, f2 = f2, f3 = f3, f4 = f4)

# Mask to use when estimating abundance and density, and maps
extrap_mask <- mask_survey_area_elev

# User-distance matrix matching the extrapolation mask
extrap_mask_userd <- userd_survey_area_elev

# `mask_model` was used for fitting, but abundance is extrapolated to `extrap_mask`
survey_area_km2 <- nrow(extrap_mask) * attr(extrap_mask, "area") / 100
```

The abundance table uses expected abundance, `E.N`, from `region.N()`. Density is abundance divided by the reporting-mask area in square kilometres.

```{r full-abundance-aic}
# To use region.N to estimate abundance on a new mask, need to replace the model's
# userdist with the new mask's userdist. Do this for all models so that we can estimate
# abundance for all models.
fitted_models_region <- lapply(fitted_models, function(model) {
  model$details$userdist <- extrap_mask_userd
  model
})

# Function to extract abundance estimates from secr's region.N
# abundance = E.N (expected abundance) or R.N (realised abundance)
extract_abundance <- function(model_name, abundance = c("E.N")) {
  rn <- region.N(fitted_models_region[[model_name]], region = extrap_mask)
  data.frame(
    modelname = model_name,
    estimate = rn[abundance, "estimate"],
    SE.estimate = rn[abundance, "SE.estimate"],
    lcl = rn[abundance, "lcl"],
    ucl = rn[abundance, "ucl"]
  )
}

# Estimate abundance for each fitted model over the survey area.
# Add density = abundance / area.
abundance_table <- lapply(names(fitted_models), extract_abundance) %>%
  bind_rows() %>%
  mutate(
    D = estimate / survey_area_km2,
    Dlcl = lcl / survey_area_km2,
    Ducl = ucl / survey_area_km2,
    CV = 100 * SE.estimate / estimate
  )

# Add AIC results and combine them with abundance and density
aic_table <- AIC(f0, f1, f2, f3, f4, criterion = "AICc") %>%
  mutate(modelname = row.names(.)) %>%
  left_join(abundance_table, by = "modelname")

# Save the combined summary table
write.csv(
  aic_table %>%
    dplyr::select(modelname, model, AIC, AICc, dAICc, AICcwt, estimate, lcl, ucl, D, Dlcl, Ducl, CV),
  file = file.path(results_dir, "namkha_advanced_results_abundance_AIC_full.csv"),
  row.names = FALSE
)

# Use the model with the lowest AICc for the density surface map
best_model_name <- aic_table$modelname[which.min(aic_table$AICc)]
best_model <- fitted_models_region[[best_model_name]]
aic_table
```

The survey-region map shows the Namkha boundary, the trap-buffer region, the gorge, and the camera locations. It is useful for checking the spatial relationship between the fitted and reported regions.

```{r full-survey-map, fig.cap="Namkha survey area, trap-buffer region, gorge, and full-survey camera locations."}
# Convert traps, the survey area boundary, and the gorge to sf objects for plotting
traps_sf <- st_as_sf(full_traps, coords = c("x", "y"), crs = my_crs)
survey_area_sf <- st_as_sf(survey_area)
gorge_sf <- st_as_sf(gorge)

# Plot the Namkha boundary, the trap-buffer region, the gorge, and the camera locations
survey_region_plot <- ggplot() +
  geom_sf(data = survey_area_sf, fill = NA, colour = "black", linewidth = 0.5) +
  geom_sf(data = st_as_sf(all_sess_masks), fill = NA, colour = "firebrick", linewidth = 0.4, linetype = 2) +
  geom_sf(data = gorge_sf, colour = "steelblue4", linewidth = 0.6) +
  geom_sf(data = traps_sf, colour = "black", size = 1.1) +
  labs(x = NULL, y = NULL) +
  theme_bw(base_size = 12) +
  theme(panel.grid = element_blank())

survey_region_plot
```

The density map is made from the lowest-AICc model. `predictDsurface()` returns density per hectare, so the values are multiplied by 10000 to display density per 100 km2.

```{r full-density-map, fig.cap="Predicted full-survey density over the elevation-constrained Namkha reporting mask."}
# User sets the model to make the map for.
# Usually set this to best_model.
map_model <- best_model

# Predict density over the Namkha survey area
pred_surface <- predictDsurface(map_model, mask = extrap_mask, cl.D = FALSE)

# Convert mask points to square cells so the predicted density is easier to plot
density_cells <- st_as_sf(data.frame(extrap_mask), coords = c("x", "y"), crs = my_crs)
density_cells <- st_buffer(
  density_cells,
  dist = attr(extrap_mask, "spacing") / 2,
  endCapStyle = "SQUARE",
  nQuadSegs = 1
)

# Convert density from animals per hectare to animals per 100 km2
density_cells$D_per_100km2 <- covariates(pred_surface)[, "D.0"] * 10000

# Plot predicted density together with the survey boundary, gorge, and trap locations
density_plot <- ggplot() +
  geom_sf(data = density_cells, aes(fill = D_per_100km2), colour = NA, alpha = 0.85) +
  geom_sf(data = survey_area_sf, fill = NA, colour = "black", linewidth = 0.5) +
  geom_sf(data = gorge_sf, colour = "steelblue4", linewidth = 0.6) +
  geom_sf(data = traps_sf, colour = "black", size = 0.8) +
  scale_fill_viridis_c(name = expression("Density / 100 km"^2), option = "D") +
  labs(x = NULL, y = NULL, title = "Predicted density") +
  theme_bw(base_size = 12) +
  theme(panel.grid = element_blank(), legend.position = "bottom")

density_plot

# Save survey and density maps
ggsave(file.path(results_dir, "namkha_advanced_survey_region_full.png"), survey_region_plot, width = 7, height = 5, dpi = 200)
ggsave(file.path(results_dir, "namkha_advanced_predicted_density_full.png"), density_plot, width = 7, height = 5, dpi = 200)

# Save the predicted density surface as a shapefile for later mapping
st_write(
  density_cells %>% dplyr::select(D_per_100km2),
  dsn = file.path(results_dir, "namkha_advanced_predicted_density_full.shp"),
  delete_layer = TRUE,
  quiet = TRUE
)
```

For covariate-effect plots we use `f4`, because it includes both the density and detection covariates used in these examples. The density plot varies ruggedness while holding any other density covariates constant.

```{r full-density-covariate-plot, fig.cap="Predicted density against terrain ruggedness for the full-survey example model."}
# User sets the model to make plot for.
# f4 includes the density and detection covariates used in these example plots.
covplot_model <- fitted_models_region[["f4"]]

# Identify which covariates appear in the density part of the chosen model
model_covs <- covplot_model$betanames
dens_covs <- sub("^D\\.", "", model_covs[grepl("^D\\.", model_covs)])
dens_covs <- intersect(dens_covs, names(covariates(extrap_mask)))
dens_covs

# User sets the density covariate to plot.
# Any other density covariates in the same model will be held constant.
focal_cov <- "std_tri"
focal_cov_full_name <- "Terrain ruggedness index"

if (!(focal_cov %in% dens_covs)) {
  stop("focal_cov must be a density covariate in chosen model.")
}

newdata <- data.frame(
  focal_values = seq(
    min(covariates(extrap_mask)[, focal_cov], na.rm = TRUE),
    max(covariates(extrap_mask)[, focal_cov], na.rm = TRUE),
    length.out = min(500, nrow(extrap_mask))
  )
)
names(newdata)[1] <- focal_cov

focal_cov_unstd <- sub("^std_", "", focal_cov)
plot_cov <- focal_cov

if (startsWith(focal_cov, "std_") && focal_cov_unstd %in% scaling_df$cov) {
  focal_mean <- scaling_df[scaling_df$cov == focal_cov_unstd, "mean"]
  focal_sd <- scaling_df[scaling_df$cov == focal_cov_unstd, "sd"]
  newdata[[focal_cov_unstd]] <- newdata[[focal_cov]] * focal_sd + focal_mean
  plot_cov <- focal_cov_unstd
}

other_dens_covs <- setdiff(dens_covs, focal_cov)
for (covname in other_dens_covs) {
  x <- covariates(extrap_mask)[, covname]
  x <- x[!is.na(x)]
  if (is.numeric(x) && length(unique(x)) > 5) {
    newdata[[covname]] <- mean(x)
  } else {
    x_unique <- unique(x)
    x_mode <- x_unique[which.max(tabulate(match(x, x_unique)))]
    newdata[[covname]] <- x_mode
  }
}

pred_mask <- subset(extrap_mask, subset = seq_len(nrow(extrap_mask)) <= nrow(newdata))
for (covname in names(newdata)) {
  covariates(pred_mask)[, covname] <- newdata[[covname]]
}

tmp <- predictDsurface(covplot_model, mask = pred_mask, cl.D = TRUE)
newdata$est <- 10000 * covariates(tmp)[, "D.0"]
newdata$lcl <- 10000 * covariates(tmp)[, "lcl.0"]
newdata$ucl <- 10000 * covariates(tmp)[, "ucl.0"]

ymax <- min(max(newdata$est * 2), max(newdata$ucl))
x_surveyed <- data.frame(xt = c(covariates(mask_model)[, plot_cov]))
x_full <- data.frame(xt = c(covariates(extrap_mask)[, plot_cov]))

pD <- newdata %>%
  ggplot(aes(x = .data[[plot_cov]], y = est)) +
  geom_line(colour = "black") +
  geom_ribbon(aes(ymin = lcl, ymax = ucl), fill = "grey70", colour = NA, alpha = 0.5) +
  geom_rug(data = x_full, inherit.aes = FALSE, aes(x = xt), sides = "b", alpha = 0.1, linewidth = 0.25) +
  geom_rug(data = x_surveyed, inherit.aes = FALSE, aes(x = xt), sides = "t", alpha = 0.1, linewidth = 0.25) +
  labs(x = focal_cov_full_name, y = bquote("Density / 100km"^2)) +
  theme_bw(base_size = 14) +
  coord_cartesian(ylim = c(0, ymax)) +
  theme(panel.grid = element_blank(), strip.background = element_rect(fill = "white"))

pD
ggsave(file.path(results_dir, "namkha_advanced_Dcovs_full.png"), pD, width = 9, height = 3.5, dpi = 300)
```

The detection plots use the same model but change `lam0_covs_vals`. For `f4`, the order is cliff then valley/stream, so `c(0, 0)` is baseline, `c(1, 0)` is cliff, and `c(0, 1)` is valley/stream.

The y-axis is encounters per day, not detection probability. This follows from the count-detector model: camera effort was measured in days, so `lambda0` and the encounter-rate curve are on the same temporal scale. The helper function calculates the encounter-rate curve over a sequence of distances and applies the selected detector-covariate values by adding the corresponding `lambda0` coefficients to the intercept on the link scale.

```{r full-detection-covariate-plots, fig.cap="Encounter-rate curves for baseline, cliff, and valley/stream detector settings in the full-survey model."}
# Identify which covariates appear in the encounter-rate part of the chosen model
lam0_covs <- sub("^lambda0\\.", "", model_covs[grepl("^lambda0\\.", model_covs)])
lam0_covs

plot_detection_setting <- function(lam0_covs_vals) {
  if (length(lam0_covs_vals) != length(lam0_covs)) {
    stop("lam0_covs_vals must be the same length as lam0_covs.")
  }

  ecd <- plot_detfunction(
    model = covplot_model,
    nd = 100,
    nbins = 25,
    lam0_covs_vals = lam0_covs_vals
  )

  ggplot() +
    geom_line(data = ecd$detfun, aes(x = d / 1000, y = e.est), linewidth = 1.1) +
    geom_ribbon(data = ecd$detfun, aes(x = d / 1000, ymin = e.lcl, ymax = e.ucl), alpha = 0.2) +
    xlab("Distance (km)") +
    ylab("Encounters per day") +
    theme_bw(base_size = 14) +
    theme(panel.grid = element_blank(), strip.background = element_rect(fill = "white"))
}

ec_baseline <- plot_detection_setting(rep(0, length(lam0_covs)))
ec_cliff <- plot_detection_setting(c(1, 0))
ec_valley_stream <- plot_detection_setting(c(0, 1))

ec_baseline
ec_cliff
ec_valley_stream

ggsave(file.path(results_dir, "namkha_advanced_Ecovs_full_baseline.png"), ec_baseline, width = 9, height = 4.5, dpi = 300)
ggsave(file.path(results_dir, "namkha_advanced_Ecovs_full_cliff.png"), ec_cliff, width = 9, height = 4.5, dpi = 300)
ggsave(file.path(results_dir, "namkha_advanced_Ecovs_full_valley_stream.png"), ec_valley_stream, width = 9, height = 4.5, dpi = 300)
```

# Shortened-survey results

The shortened-survey workflow mirrors the full-survey workflow. The main differences are the fitted model list, the traps object used for plotting, and the output directory. The same reporting mask is used, so full and shortened results are reported over the same spatial scale.

```{r short-results}
results_dir <- "namkha_advanced/output/results/short"
dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)

load("namkha_advanced/output/namkha_advanced_secr_inputs_capthist.RData")
load("namkha_advanced/output/namkha_advanced_secr_inputs_mask.RData")
load("namkha_advanced/output/namkha_advanced_fitted_models_short.RData")

fitted_models <- list(s0 = s0, s1 = s1, s2 = s2, s3 = s3, s4 = s4)
extrap_mask <- mask_survey_area_elev
extrap_mask_userd <- userd_survey_area_elev
survey_area_km2 <- nrow(extrap_mask) * attr(extrap_mask, "area") / 100

fitted_models_region <- lapply(fitted_models, function(model) {
  model$details$userdist <- extrap_mask_userd
  model
})

extract_abundance <- function(model_name, abundance = c("E.N")) {
  rn <- region.N(fitted_models_region[[model_name]], region = extrap_mask)
  data.frame(
    modelname = model_name,
    estimate = rn[abundance, "estimate"],
    SE.estimate = rn[abundance, "SE.estimate"],
    lcl = rn[abundance, "lcl"],
    ucl = rn[abundance, "ucl"]
  )
}

abundance_table <- lapply(names(fitted_models), extract_abundance) %>%
  bind_rows() %>%
  mutate(
    D = estimate / survey_area_km2,
    Dlcl = lcl / survey_area_km2,
    Ducl = ucl / survey_area_km2,
    CV = 100 * SE.estimate / estimate
  )

aic_table <- AIC(s0, s1, s2, s3, s4, criterion = "AICc") %>%
  mutate(modelname = row.names(.)) %>%
  left_join(abundance_table, by = "modelname")

write.csv(
  aic_table %>%
    dplyr::select(modelname, model, AIC, AICc, dAICc, AICcwt, estimate, lcl, ucl, D, Dlcl, Ducl, CV),
  file = file.path(results_dir, "namkha_advanced_results_abundance_AIC_short.csv"),
  row.names = FALSE
)

best_model_name <- aic_table$modelname[which.min(aic_table$AICc)]
best_model <- fitted_models[[best_model_name]]
aic_table
```

The remaining shortened-survey maps and covariate plots use the same code pattern as the full-survey section, with `short_traps`, `s4`, and the `short` output filenames. The script `05b_model_results_short.R` contains the complete plotting code for running this as a standalone script.

# Multisession results

Multisession reporting differs because `region.N()` returns a result for each session. The reporting `userdist` also needs to be a named list, matching the names of `traps_ms`.

The model `ms5`, with `D ~ session`, is omitted from this reporting workflow because session as a density covariate makes extrapolation to a non-session reporting mask awkward. The reported model set uses the shared-parameter models `ms0` to `ms4`.

```{r ms-abundance-aic}
results_dir <- "namkha_advanced/output/results/ms"
dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)

load("namkha_advanced/output/namkha_advanced_secr_inputs_capthist.RData")
load("namkha_advanced/output/namkha_advanced_secr_inputs_mask.RData")
load("namkha_advanced/output/namkha_advanced_secr_inputs_msmask.RData")
load("namkha_advanced/output/namkha_advanced_fitted_models_ms.RData")

# Omitting ms5 which has D ~ session
fitted_models <- list(ms0 = ms0, ms1 = ms1, ms2 = ms2, ms3 = ms3, ms4 = ms4)

extrap_mask <- mask_survey_area_elev
extrap_mask_userd <- userd_survey_area_elev
survey_area_km2 <- nrow(extrap_mask) * attr(extrap_mask, "area") / 100

fitted_models_region <- lapply(fitted_models, function(model) {
  model$details$userdist <- setNames(
    replicate(length(traps_ms), extrap_mask_userd, simplify = FALSE),
    names(traps_ms)
  )
  model
})

extract_abundance_ms <- function(model_name, abundance = c("E.N")) {
  rn <- region.N(fitted_models_region[[model_name]], region = extrap_mask)
  bind_rows(lapply(seq_along(rn), function(i) {
    data.frame(
      session = names(rn)[i],
      modelname = model_name,
      estimate = rn[[i]][abundance, "estimate"],
      SE.estimate = rn[[i]][abundance, "SE.estimate"],
      lcl = rn[[i]][abundance, "lcl"],
      ucl = rn[[i]][abundance, "ucl"]
    )
  }))
}

abundance_table <- lapply(names(fitted_models), extract_abundance_ms) %>%
  bind_rows() %>%
  mutate(
    D = estimate / survey_area_km2,
    Dlcl = lcl / survey_area_km2,
    Ducl = ucl / survey_area_km2,
    CV = 100 * SE.estimate / estimate
  )

aic_table <- AIC(ms0, ms1, ms2, ms3, ms4, criterion = "AICc") %>%
  mutate(modelname = row.names(.))

write.csv(
  left_join(abundance_table, aic_table, by = "modelname") %>%
    dplyr::select(session, modelname, model, AIC, AICc, dAICc, AICcwt, estimate, lcl, ucl, D, Dlcl, Ducl, CV),
  file = file.path(results_dir, "namkha_advanced_results_abundance_AIC_ms.csv"),
  row.names = FALSE
)

best_model_name <- aic_table$modelname[which.min(aic_table$AICc)]
best_model <- fitted_models[[best_model_name]]
left_join(abundance_table, aic_table, by = "modelname")
```

The multisession density map uses the lowest-AICc reported multisession model. `predictDsurface()` returns one density surface per session, even when the model shares parameters across sessions, so the script explicitly uses the first session's density column for mapping.

```{r ms-density-map, fig.cap="Predicted multisession density over the elevation-constrained Namkha reporting mask, using the first session surface."}
# Convert traps, the survey area boundary, and the gorge to sf objects for plotting
traps_sf <- st_as_sf(full_traps, coords = c("x", "y"), crs = my_crs)
survey_area_sf <- st_as_sf(survey_area)
gorge_sf <- st_as_sf(gorge)

survey_region_plot <- ggplot() +
  geom_sf(data = survey_area_sf, fill = NA, colour = "black", linewidth = 0.5) +
  geom_sf(data = st_as_sf(all_sess_masks), fill = NA, colour = "firebrick", linewidth = 0.4, linetype = 2) +
  geom_sf(data = gorge_sf, colour = "steelblue4", linewidth = 0.6) +
  geom_sf(data = traps_sf, colour = "black", size = 1.1) +
  labs(x = NULL, y = NULL) +
  theme_bw(base_size = 12) +
  theme(panel.grid = element_blank())

ggsave(file.path(results_dir, "namkha_advanced_survey_region_ms.png"), survey_region_plot, width = 7, height = 5, dpi = 200)

# User sets the model to make the map for.
# Usually set this to best_model.
map_model <- best_model

# The multisession model predicts one density surface per session, even when
# predicting over a single reporting mask. Use the first session's surface here.
map_session <- names(traps_ms)[1]

# Predict density over the Namkha survey area
pred_surface <- predictDsurface(map_model, mask = extrap_mask, cl.D = FALSE)
density_cols <- which(names(covariates(pred_surface)) == "D.0")

if (length(density_cols) < 1) {
  stop("No predicted density columns found in pred_surface.")
}

density_cells <- st_as_sf(data.frame(extrap_mask), coords = c("x", "y"), crs = my_crs)
density_cells <- st_buffer(
  density_cells,
  dist = attr(extrap_mask, "spacing") / 2,
  endCapStyle = "SQUARE",
  nQuadSegs = 1
)

density_cells$D_per_100km2 <- covariates(pred_surface)[, density_cols[1]] * 10000

density_plot <- ggplot() +
  geom_sf(data = density_cells, aes(fill = D_per_100km2), colour = NA, alpha = 0.85) +
  geom_sf(data = survey_area_sf, fill = NA, colour = "black", linewidth = 0.5) +
  geom_sf(data = gorge_sf, colour = "steelblue4", linewidth = 0.6) +
  geom_sf(data = traps_sf, colour = "black", size = 0.8) +
  scale_fill_viridis_c(name = expression("Density / 100 km"^2), option = "D") +
  labs(x = NULL, y = NULL, title = paste("Predicted density from session", map_session)) +
  theme_bw(base_size = 12) +
  theme(panel.grid = element_blank(), legend.position = "bottom")

density_plot

ggsave(file.path(results_dir, "namkha_advanced_predicted_density_ms.png"), density_plot, width = 7, height = 5, dpi = 200)

st_write(
  density_cells %>% dplyr::select(D_per_100km2),
  dsn = file.path(results_dir, "namkha_advanced_predicted_density_ms.shp"),
  delete_layer = TRUE,
  quiet = TRUE
)
```

The multisession density and detection covariate plots follow the same pattern as the full-survey examples, using `covplot_model <- fitted_models[["ms4"]]`. As with the density map, duplicate `D.0`, `lcl.0`, and `ucl.0` columns can be returned; the script explicitly extracts the first matching set.

```{r ms-covariate-column-check}
# User sets the model to make plot for.
# ms4 includes the density and detection covariates used in these example plots.
covplot_model <- fitted_models[["ms4"]]

# Identify which covariates appear in the density part of the chosen model.
# Session effects are model parameters rather than mask covariates, so only
# retain density covariates that are present in the extrapolation mask.
model_covs <- covplot_model$betanames
dens_covs <- sub("^D\\.", "", model_covs[grepl("^D\\.", model_covs)])
dens_covs <- intersect(dens_covs, names(covariates(extrap_mask)))
dens_covs

lam0_covs <- sub("^lambda0\\.", "", model_covs[grepl("^lambda0\\.", model_covs)])
lam0_covs
```

# What comes next

These outputs support comparison of the main advanced approaches: using the full survey, shortening the survey window, and splitting the survey into temporal sessions. In an applied analysis, the same reporting pattern can be extended to additional candidate models, but the fitting mask, reporting mask, and matching `userdist` object should always remain aligned.
