---
title: Simple SECR workflow
subtitle: Namkha basic tutorial
author: |
  | Snow leopard SECR workflow workshop
  | Ian Durbach and Cornelia Oedekoven
institute: |
  | Centre for Research into Ecological & Environmental Modelling
  | University of St Andrews
fontsize: 10pt
format:
  beamer:
    theme: metropolis
    incremental: false
    slide-level: 2
    pdf-engine: xelatex
    syntax-highlighting: tango
classoption: t
urlcolor: teal
execute:
  echo: true
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(knitr)
library(secr)
library(dplyr)
library(ggplot2)
library(MASS)
source("../code//more-utilities.R")

knitr::opts_knit$set(global.par = TRUE)
knitr::opts_chunk$set(fig.path = "code-and-figures/")

def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(
  chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    chunk_size <- if (is.null(options$size)) "normalsize" else options$size
    ifelse(
      chunk_size != "normalsize",
      paste0("\n \\", chunk_size, "\n\n", x, "\n\n \\normalsize"),
      x
    )
  }
)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines) == 1) {
    if (length(x) > lines) {
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})

load("../output/namkha_basic_secr_inputs_capthist.RData")
load("../output/namkha_basic_secr_inputs_mask.RData")
load("../output/namkha_basic_fitted_models.RData")
extrap_mask <- mask_survey_area
```

## Session 2 objectives

Main goal: Demonstrate the workflow with a simple example where everything "works":

1. Create a single-session `capthist` object from camera and detection files
2. Build habitat masks for model fitting and reporting
3. Fit a set of `secr` models
4. Turn fitted models into density, abundance, maps, and checks

\vspace{0.4cm}

The emphasis is the workflow: what each object means and how the scripts connect.

## The simple workflow

Each script saves the main object needed by the next step:

\begin{tikzpicture}[remember picture, overlay]
\node[xshift=-4.5cm, yshift=-0.8cm, draw=black, fill=blue!10,
      fill opacity = 0.3, text opacity = 1, shape=rectangle,
      rounded corners=1.5] at (current page){
      \begin{minipage}[t][3.5cm][t]{0.15\textwidth}
       \textbf{01}\\
       camera file\\
       detection file\\
       $\Rightarrow$ \\
       \texttt{capthist}
      \end{minipage}};
\node[xshift=-2.25cm, yshift=-0.8cm, draw=black, fill=blue!10,
      fill opacity = 0.3, text opacity = 1, shape=rectangle,
      rounded corners=1.5] at (current page){
      \begin{minipage}[t][3.5cm][t]{0.15\textwidth}
       \textbf{02}\\
       survey area\\
       buffer\\
       covariates\\
       $\Rightarrow$ \\
       \texttt{mask}
      \end{minipage}};
\node[xshift=0cm, yshift=-0.8cm, draw=black, fill=red!20,
      fill opacity = 0.3, text opacity = 1, shape=rectangle,
      rounded corners=1.5] at (current page){
      \begin{minipage}[t][3.5cm][t]{0.17\textwidth}
       \textbf{03}\\
       (crop for \\ 
       multisession \\
       model) \\
       $\Rightarrow$ \\
       \texttt{mask (ms)}
      \end{minipage}};
\node[xshift=2.25cm, yshift=-0.8cm, draw=black, fill=blue!10,
      fill opacity = 0.3, text opacity = 1, shape=rectangle,
      rounded corners=1.5] at (current page){
      \begin{minipage}[t][3.5cm][t]{0.15\textwidth}
       \textbf{04}\\
       $D$ model\\
       $\lambda_0$/$g_0$ model\\
       $\Rightarrow$ \\
       \texttt{secr.fit}
      \end{minipage}};
\node[xshift=4.5cm, yshift=-0.8cm, draw=black, fill=blue!10,
      fill opacity = 0.3, text opacity = 1, shape=rectangle,
      rounded corners=1.5] at (current page){
      \begin{minipage}[t][3.5cm][t]{0.15\textwidth}
       \textbf{05}\\
       AICc\\
       abundance\\
       density\\
       maps
      \end{minipage}};
\end{tikzpicture}

## Capture histories: script `01_make_capthist.R`

`01_make_capthist.R` creates a `capthist` object for `secr`. This contains:

1. the **trapfile**: detector locations and related covariates (a `traps` object)
2. the **encounter histories**: recorded encounters of animals at detectors (a data frame)

Tasks involved:

- read and clean camera records
- read detection records
- encode effort as trap usage
- attach trap-level covariates
- create and check a `capthist`
- save the object for later scripts

## Capture histories: camera data to `traps`

Set file paths and CRS used by the workflow

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
my_crs <- "+proj=utm +zone=44 +datum=WGS84 +units=m +no_defs"

traps_file <- "namkha_basic/data/traps.csv"
detections_file <- "namkha_basic/data/detections.csv"

output_file <- "namkha_basic/output/namkha_basic_secr_inputs_capthist.RData"
fig_dir <- "namkha_basic/output/fig"
```
\normalsize

Read in the camera location file (intentionally clean but see later!)

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
cameras <- read.csv(traps_file, stringsAsFactors = FALSE)
```
\normalsize

## Capture histories: camera data to `traps`

\footnotesize
```{r}
#| eval: true
#| echo: false
#| output: true
cameras <- read.csv("../data/traps.csv", stringsAsFactors = FALSE)
head(cameras, 8)
```
\normalsize

## Capture histories: camera data to `traps`

Drop malfunctioning or lost cameras

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
cameras <- cameras |> filter(Note == "Functional") 
```
\normalsize

Ensure correct date format

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
cameras <- cameras |>
  mutate(
    start_date = as.Date(start_date, format = "%d/%m/%Y"),
    end_date   = as.Date(end_date,   format = "%d/%m/%Y")
  )
```
\normalsize

Define the overall survey period (here from active cameras)

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
survey_start <- min(cameras$start_date, na.rm = TRUE)
survey_end <- max(cameras$end_date, na.rm = TRUE)
```
\normalsize

## Capture histories: camera data to `traps`

Create the `secr` `traps` object

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
traps <- read.traps(
  data = traps_df |> dplyr::select(trapID, x, y, effort),
  detector = "count",
  trapID = "trapID",
  binary.usage = FALSE
)
```
\normalsize

Add covariates that might influence detectability of animals

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
covariates(traps)$Topography    <- traps_df$Topography
covariates(traps)$cliff         <- traps_df$cliff
covariates(traps)$hill          <- traps_df$hill
covariates(traps)$valley_stream <- traps_df$valley_stream
covariates(traps)$Water         <- traps_df$Water
covariates(traps)$Altitude      <- traps_df$Altitude
```
\normalsize

## Capture histories: detection data to `capthist`

Read in the detection data (clean; see later for more realistic examples!)

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
capts_raw <- read.csv(detections_file, stringsAsFactors = FALSE)
```
\normalsize

\vspace{.6em}
\footnotesize
```{r}
#| eval: true
#| echo: false
#| output: true
capts_raw <- read.csv("../data/detections.csv", stringsAsFactors = FALSE)
head(capts_raw, 12)
```
\normalsize

## Capture histories: detection data to `capthist`

Only keep detections in the survey period

\vspace{.6em}
\footnotesize
```{r}
#| code-line-numbers: "1|2-3"
#| eval: false
#| echo: true
capts_raw <- capts_raw |>
  mutate(date = as.Date(date, format = "%m/%d/%Y")) |>
  filter(date >= survey_start, date <= survey_end)
```
\normalsize

Keep variables used by `secr`

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
capts <- capts_raw |>
  dplyr::select(session, animalID, occasion, trapID, sex)
```
\normalsize

Create and inspect the `secr` `capthist` object:

\vspace{.6em}
\footnotesize
```{r}
#| code-line-numbers: "1|2|3"
#| eval: false
#| echo: true
ch <- make.capthist(captures = capts, traps = traps)
verify(ch)
summary(ch, terse = TRUE)
```
\normalsize

## Capture histories: descriptive checks

Before fitting models, check that effort and detections are coherent:

\begin{columns}[T] % align at top
  \begin{column}{0.7\textwidth}
    \centering
    \includegraphics[width=\linewidth]{../output/fig/namkha_capture_history_summary.png}
  \end{column}

  \begin{column}{0.3\textwidth}
  \vspace{2em}
    \begin{itemize}
      \item do detections occur during active camera periods?
      \item new individuals distributed over cameras?
      \item do new individuals keep appearing late in the survey?
    \end{itemize}
  \end{column}
\end{columns}

## Habitat masks: script `02_make_masks.R`

`02_make_masks.R` creates habitat masks (class `mask`) for fitting and prediction.

\alert{Key idea:} 

- Different masks for model fitting and model prediction
- Model fitting mask needs to include all AC locations animals could be detected from
- Often want to extrapolate to wider area
- *But* where these two masks intersect, useful to share mask points

\alert{Implementation:} 

- Create polygons for (1) close to detectors; (2) extrapolation region
- Find union of these two polygons
- Make mask (points) in this region
- Subset this to form model fitting and prediction masks

## Habitat masks: script 02

Tasks involved:

- read the Namkha boundary (extrapolation region)
- build a trap buffer 
- create a regular grid of mask points
- add spatial covariates to the mask
- save separate masks for fitting and reporting

## Habitat masks: define the region

The full region is the union of the survey boundary and the trap buffer:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
mask_spacing <- 1500
mask_buffer  <- 25000
```
\normalsize

\alert{Aside:} How to choose buffer and spacing?

- Get a rough estimate of $\sigma$ from capthist
- Rough guesses for buffer = 4$\sigma$, spacing = 0.6$\sigma$
- Bigger buffers and small spacings are safer but increase run times

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
qnd <- autoini(capthist = ch, mask = make.mask(traps = traps, 
                                               buffer = 75000, 
                                               spacing = 1000))
qnd$sigma * 4    # buffer
qnd$sigma * 0.6  # spacing
```
\normalsize

\vspace{-.6em}
\footnotesize
```{r}
#| eval: true
#| echo: false
#| output: true
qnd <- autoini(capthist = ch, 
               mask = make.mask(traps = traps, 
                                buffer = 75000, 
                                spacing = 1000))
cat("Rough buffer: ", qnd$sigma * 4)    
cat("Rough spacing: ", qnd$sigma * 0.6)  
```
\normalsize

## Habitat masks: define the region

Read in extrapolation region polygon

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true

survey_area <- st_read("namkha_basic/data/
                       survey_area/Namkha_RM.shp",
                       quiet = TRUE) |>
  st_zm() |>
  st_transform(my_crs) |>
  st_geometry()
```
\normalsize

Make traps buffer polygon

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true

traps_buffer <- traps_sf %>%
  st_buffer(mask_buffer) %>%
  st_union() %>% st_geometry()
```
\normalsize

Full region is the union of the two polygons

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true

full_region <- st_union(survey_area, traps_buffer)
```
\normalsize

## Habitat masks: create the mask

Build a regular grid

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
mask_grid <- st_make_grid(
  # add a small buffer so don't lose points on boundary
  st_buffer(full_region, 10000),
  cellsize = c(mask_spacing, mask_spacing),
  what = "centers"
)
```
\normalsize

Keep points inside the region

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
mask_points <- mask_grid[full_region] |>
  st_as_sf() |>
  cbind(st_coordinates(.)) |>
  st_drop_geometry() |>
  dplyr::select(x = X, y = Y)
```
\normalsize

Create and inspect the `secr` `mask` object

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
mask <- read.mask(data = mask_points, spacing = mask_spacing)
```
\normalsize

## Habitat masks: add spatial covariates

Density covariates added to mask because they affect possible AC locations. 

Easiest to read in as raster or shape file

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
tri <- rast("namkha_basic/data/spatial_covs/TRI_Namkha.tif")
# remember to project to same CRS as mask
tri <- project(tri, y = my_crs)
names(tri) <- "tri"
mask <- addCovariates(mask, tri)
```
\normalsize

May need some preprocessing (more later on covariates)

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
covariates(mask)$d2hydro <- as.numeric(st_distance(mask_sf, hydro))
```
\normalsize

- `mask_model` is used for `secr.fit()`
- `mask_survey_area` is used for abundance and reporting
- these can but don't have to differ

## Model fitting: script 04

`04_model_fitting.R` fits a set of candidate models.

- start with the null model
- add density covariates such as `std_tri` or `std_d2hydro`
- add detector covariates such as `valley_stream`
- use AIC for model selection

## Model fitting: adding covariates

Fitting the null model:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
m0 <- secr.fit(ch,
  detectfn = "HHN",
  mask = mask_model,
  model = list(D ~ 1, lambda0 ~ 1, sigma ~ 1)
)
```
\normalsize

`m4` has covariates on density and detection

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
m4 <- secr.fit(ch,
  detectfn = "HHN",
  mask = mask_model,
  model = list(D ~ std_tri, lambda0 ~ valley_stream, sigma ~ 1),
  start = m0
)
```
\normalsize

## Model fitting: adding covariates

Use AICc or AIC to compare the candidate set after fitting

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
AIC(m0, m1, m2, m3, m4, m5, criterion = "AICc")
```

```{r}
#| eval: true
#| echo: false
#| output: true
AIC(m0, m1, m2, m3, m4, m5, criterion = "AICc")
```
\normalsize

## Interpreting the null model -- m~0~

Type the name of the model for a model summary

\vspace{.6em}
```{r,eval=TRUE, echo=FALSE, size="scriptsize", output.lines=15}
m0 # 1 of 3
```

## Interpreting the null model -- m~0~

Type the name of the model for a model summary

\vspace{.6em}
\footnotesize
```{r,eval=TRUE, echo=FALSE, size="scriptsize", output.lines=16:31}
m0 # 1 of 3
```

## Interpreting the null model -- m~0~

Type the name of the model for a model summary

\vspace{.6em}
\footnotesize
```{r,eval=TRUE, echo=FALSE, size="scriptsize", output.lines=32:49}
m0 # 1 of 3
```

## Interpreting the null model -- m~0~

- `coef()` for model coefficients on the link scale

\vspace{.6em}
\footnotesize
```{r,eval=TRUE, echo=TRUE, size="scriptsize"}
coef(m0) 
```

- `predict()` for model coefficients on the response scale

\vspace{.6em}
\footnotesize
```{r,eval=TRUE, echo=TRUE, size="scriptsize"}
predict(m0) 
```

## Interpreting a covariate model -- m~4~

- `coef()` for model coefficients on the link scale

\vspace{.6em}
\footnotesize
```{r,eval=TRUE, echo=TRUE, size="scriptsize"}
coef(m4) 
```

- `predict()` for model coefficients on the response scale

\vspace{.6em}
\footnotesize
```{r,eval=TRUE, echo=TRUE, size="scriptsize"}
predict(m4) 
```

## `secr` workflow

\begin{tikzpicture}[remember picture, overlay]
\node[xshift=-0.5cm, yshift=-0.5cm] at (current page){
\includegraphics[height=1.6in]{inputs/secr}};
\end{tikzpicture}

\begin{tikzpicture}[remember picture, overlay]
\node[xshift=-4.38cm, yshift=-0.2cm, draw=black,fill=blue!10,
      fill opacity = 0.3, text opacity = 1, shape=rectangle, 
      rounded corners=1.5] at (current page){
      \begin{minipage}[t][4.4cm][t]{0.19\textwidth}
       {\textbf{Pre-\texttt{secr}}}
      \end{minipage}};
\node[xshift=-1.7cm, yshift=-0.2cm, draw=black,fill=blue!10,
      fill opacity = 0.3, text opacity = 1, shape=rectangle, 
      rounded corners=1.5] at (current page){
      \begin{minipage}[t][4.4cm][t]{0.26\textwidth}
       {\textbf{\texttt{secr} formatting}}
      \end{minipage}};
\node[xshift=1.2cm, yshift=-0.2cm, draw=black,fill=blue!10,
      fill opacity = 0.3, text opacity = 1, shape=rectangle, 
      rounded corners=1.5] at (current page){
      \begin{minipage}[t][4.4cm][t]{0.23\textwidth}
       {\textbf{Model fitting}}
      \end{minipage}};      
\node[xshift=3.67cm, yshift=-0.2cm, draw=black,fill=blue!10,
      fill opacity = 0.3, text opacity = 1, shape=rectangle, 
      rounded corners=1.5] at (current page){
      \begin{minipage}[t][4.4cm][t]{0.18\textwidth}
       {\textbf{Inference}}
      \end{minipage}};      
\end{tikzpicture}

## Post processing functions in `secr`

A bunch of helper functions for working with model output:

- extract link-scale estimates & associated uncertainty: `coef()`
- extract real-scale estimates & associated uncertainty: `predict()`
- compute the population size for the area: `region.N()`
- plot detection function based on estimates: `plot()`
- plot estimated AC locations: `fxi.contour()`
- many more 

## Estimation and reporting: script 05

`05_model_results.R` turns fitted models into common outputs.

Tasks involved: 

- compare models with AICc
- summarise abundance and density over `mask_survey_area`
- choose the best model for mapping by AICc
- map the survey region and predicted density
- inspect covariate effects on density and detection

## Estimation and reporting: build the summary table

Choose models you want to report results for

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
fitted_models <- list(m0 = m0, m1 = m1, m2 = m2,
                      m3 = m3, m4 = m4, m5 = m5)
```
\normalsize

Choose area you want to report results for:

\footnotesize
```{r}
#| eval: false
#| echo: true
extrap_mask <- mask_survey_area
```
\normalsize

Extracting abundance for all chosen models (`extract_abundance` is a wrapper function for `secr`'s `region.N`):

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
abundance_table <- lapply(names(fitted_models), 
                          extract_abundance, 
                          abundance = c("E.N")) 
```
\normalsize

\alert{Note:} `abundance = c("E.N")` for expected abundance, `abundance = c("R.N")` for realised abundance

## Estimation and reporting: build the summary table

Reshape into nice format:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
abundance_table <- abundance_table |> bind_rows() |>
  mutate(D = estimate / survey_area_km2,
         Dlcl = lcl / survey_area_km2, Ducl = ucl / survey_area_km2,
         CV = 100 * SE.estimate / estimate)
```
\normalsize

\vspace{.6em}
\scriptsize
```{r}
#| eval: true
#| echo: false
#| output: true

extrap_mask <- mask_survey_area
survey_area_km2 <- nrow(extrap_mask) * attr(extrap_mask, "area") / 100

extract_abundance = function(model_name, abundance = c("E.N")) {
  rn <- region.N(fitted_models[[model_name]], region = extrap_mask)
  data.frame(
    modelname = model_name,
    estimate = rn[abundance, "estimate"],
    SE.estimate = rn[abundance, "SE.estimate"],
    lcl = rn[abundance, "lcl"],
    ucl = rn[abundance, "ucl"]
  )
}

fitted_models <- list(m0 = m0, m1 = m1, m2 = m2,
                      m3 = m3, m4 = m4, m5 = m5)

abundance_table <- lapply(names(fitted_models), 
                          extract_abundance, 
                          abundance = c("E.N")) 

abundance_table |>
  bind_rows() |>
  mutate(
    D = estimate / survey_area_km2,
    Dlcl = lcl / survey_area_km2, Ducl = ucl / survey_area_km2,
    CV = 100 * SE.estimate / estimate
  )
```
\normalsize

## Estimation and reporting: abundance and density

AIC's for chosen models:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
aic_table <- AIC(m0, m1, m2, m3, m4, m5, criterion = "AICc") |>
  mutate(modelname = row.names(.)) |>
  left_join(abundance_table, by = "modelname")
```
\normalsize

Best model is one with lowest AIC:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
best_model_name <- aic_table$modelname[which.min(aic_table$AICc)]
best_model <- fitted_models[[best_model_name]]
```
\normalsize

Save results:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
write.csv(aic_table |>
            select(modelname, model, AIC, AICc, dAICc, AICcwt, 
                   estimate, lcl, ucl, D, Dlcl, Ducl, CV),
          file = "output/namkha_basic_results_abundance_AIC.csv",
          row.names = FALSE)
```
\normalsize

## Estimation and reporting: survey region

\begin{center}
\includegraphics[height=2.45in]{../output/results/namkha_basic_survey_region.png}
\end{center}

Survey boundary, camera locations, and the trap-buffer region used in the workflow.

## Estimation and reporting: predicted density

Predicted density at each point in the extrapolation mask:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
pred_surface <- predictDsurface(best_model, 
                                mask = extrap_mask, 
                                cl.D = FALSE)
```
\normalsize

Create grid cells centered on mask points for plotting:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
density_cells <- st_as_sf(data.frame(extrap_mask), 
                          coords = c("x", "y"), crs = my_crs)
density_cells <- st_buffer(
  density_cells,
  dist = attr(extrap_mask, "spacing") / 2,
  endCapStyle = "SQUARE",
  nQuadSegs = 1
)
```
\normalsize

Add in the densities per 100 km$^2$:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
density_cells$D_per_100km2 <- covariates(pred_surface)[, "D.0"] * 10000
```
\normalsize

## Estimation and reporting: predicted density

\begin{center}
\includegraphics[height=2.45in]{../output/results/namkha_basic_predicted_density.png}
\end{center}

## Estimation and reporting: density covariate effects

To show how density changes with one focal covariate, choose a model and covariate
 
\vspace{.6em}
\footnotesize
```{r}
#| eval: true
#| echo: true
covplot_model <- m5
focal_cov <- "std_tri"
focal_cov_full_name <- "Terrain ruggedness index"
```
\normalsize

Would usually choose the AIC-best model but here using `m5` which has two density covariates and two detection covariates

\vspace{.6em}
\scriptsize
```{r}
#| eval: true
#| echo: false
#| output: true
coef(m5)
```
\normalsize

## Estimation and reporting: density covariate effects

Create a range of potential values for the focal covariate:

\vspace{.6em}
\footnotesize
```{r}
#| eval: true
#| echo: true
newdata <- data.frame(
  focal_values = seq(
    min(covariates(extrap_mask)[, focal_cov], na.rm = TRUE),
    max(covariates(extrap_mask)[, focal_cov], na.rm = TRUE),
    length.out = min(500, nrow(extrap_mask))
  )
)
names(newdata)[1] <- focal_cov
```
\normalsize

Other covariates are set to their mean or mode (not shown here)

## Estimation and reporting: density covariate effects

Predicting densities at new covariate values is a bit fiddly. 

- Use a temporary mask so `predictDsurface` can evaluate the density model.

\vspace{.6em}
\footnotesize
```{r}
#| eval: true
#| echo: true
pred_mask <- subset(extrap_mask, subset = 1:nrow(newdata))
```
\normalsize

- Replace covariate values with the ones we want

\vspace{.6em}
\footnotesize
```{r}
#| eval: true
#| echo: true
for (covname in names(newdata)) {
  covariates(pred_mask)[, covname] <- newdata[[covname]]
}
```
\normalsize

- Now predict density 

\vspace{.6em}
\footnotesize
```{r}
#| eval: true
#| echo: true
tmp <- predictDsurface(covplot_model, mask = pred_mask, cl.D = TRUE)
```
\normalsize

## Estimation and reporting: density covariate effects

- Extract observed covariate values on mask and extrapolation region to assess covariate coverage and include this on plot

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
x_surveyed <- data.frame(xt = c(covariates(mask_model)[, plot_cov]))
x_full <- data.frame(xt = c(covariates(extrap_mask)[, plot_cov]))
```
\normalsize

- Finally plot: 

\begin{center}
\includegraphics[width=.8\textwidth]{../output/results/namkha_Dcovs.png}
\end{center}

## Estimation and reporting: encounter rate covariate effects

Identify which covariates appear in the encounter-rate part of the chosen model

\vspace{.6em}
\footnotesize
```{r}
#| eval: true
#| echo: true
#| output: true
model_covs <- covplot_model$betanames
lam0_covs <- sub("^lambda0\\.", "", 
                 model_covs[grepl("^lambda0\\.", model_covs)])
lam0_covs
```

Set possible values for any $\lambda_0$ covariates **in same order as** `lam0_covs`

\vspace{.6em}
\footnotesize
```{r}
#| eval: true
#| echo: true
lam0_covs_vals <- c(1, 0) # for cliff
lam0_covs_vals <- c(0, 1) # for valley/stream 
lam0_covs_vals <- c(0, 0) # for baseline (other)
```
\normalsize

## Estimation and reporting: encounter rate covariate effects

Plot for baseline:

\begin{center}
\includegraphics[width=0.5\textwidth]{../output/results/namkha_Ecovs_base.png}
\end{center}

Plot for cliff:

\begin{center}
\includegraphics[width=0.5\textwidth]{../output/results/namkha_Ecovs_cliff.png}
\end{center}

## Workflow recap

\vspace{2em}

1. Read in camera and detection csv's and create `capthist` \vspace{1em}
2. Build masks for model fitting and reporting (keep these concepts separate) \vspace{1em}
3. Fit candidate `secr` models \vspace{1em}
4. Produce interpretable output: density, abundance, maps, covariate effects

## What did we miss?

\vspace{2em}

1. Survey period too long for population closure \vspace{1em}
2. Barriers to movement \vspace{1em}
3. Other covariates  \vspace{1em}
4. Sex-dependent parameters \vspace{1em}
5. Perfectly clean data doesn't exist \vspace{1em}
6. Probably lots more!

\vspace{1em}

\Large
\alert{Some of these are covered in the next sections}
\normalsize