---
title: Simple SECR workflow
subtitle: Introduction
author: |
  | Snow leopard SECR workflow workshop
  | Ian Durbach and Cornelia Oedekoven
institute: |
  | Centre for Research into Ecological & Environmental Modelling
  | University of St Andrews
fontsize: 10pt
format:
  beamer:
    theme: metropolis
    incremental: false
    slide-level: 2
    pdf-engine: xelatex
    syntax-highlighting: tango
classoption: t
urlcolor: teal
execute:
  echo: true
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(knitr)
library(secr)
library(dplyr)
library(ggplot2)
library(MASS)
source("../../namkha_basic/code//more-utilities.R")

knitr::opts_knit$set(global.par = TRUE)
knitr::opts_chunk$set(fig.path = "code-and-figures/")

def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(
  chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    chunk_size <- if (is.null(options$size)) "normalsize" else options$size
    ifelse(
      chunk_size != "normalsize",
      paste0("\n \\", chunk_size, "\n\n", x, "\n\n \\normalsize"),
      x
    )
  }
)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines) == 1) {
    if (length(x) > lines) {
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})

load("../../namkha_basic/output/namkha_basic_secr_inputs_capthist.RData")
load("../../namkha_basic/output/namkha_basic_secr_inputs_mask.RData")
load("../../namkha_basic/output/namkha_basic_fitted_models.RData")
traps_df <- read.csv("../../namkha_basic/data/traps.csv")
caps_df <- read.csv("../../namkha_basic/data/detections.csv")
extrap_mask <- mask_survey_area
```

## Welcome

A standardized workflow for designing and analyzing camera trap surveys of large cats with spatially explicit capture-recapture (SCR).

-   Session 1: Concepts
-   Session 2: Simple workflow
-   Session 3: Adding complexity
-   Session 4: Automatic checks of data integrity
-   Session 5: Wrap up

## Welcome

A standardized workflow for designing and analyzing camera trap surveys of large cats with spatially explicit capture-recapture (SCR).

-   Session 1: Concepts
-   Session 2: Simple workflow
-   Session 3: Adding complexity
-   Session 4: Automatic checks of data integrity
-   Session 5: Wrap up

What we expect from participants:

-   Some knowledge on statistics
-   Know how to work with R and RStudio


## Ready for the practicals

Data: 

1. Download from XXXXXXXXX
2. Unzip into dedicated `secr` workshop folder

Software required:

1.  R: <https://cran.rstudio.com/>
2.  RStudio: <https://posit.co/download/rstudio-desktop>
3.  R packages (and dependencies): \vspace{.6em}

\centering
\begin{columns}[T] % align at top
  \begin{column}{0.25\textwidth}
  \end{column}
  \begin{column}{0.25\textwidth}
  \vspace{1em}
    \begin{itemize}
     \item dplyr
     \item elevatr
     \item gdistance
     \item ggplot2
     \item lubridate
     \item MASS
    \end{itemize}
  \end{column}

  \begin{column}{0.25\textwidth}
  \vspace{1em}
    \begin{itemize}
      \item patchwork
      \item secr
      \item sf
      \item stringr
      \item tidyr
      \item terra
    \end{itemize}
  \end{column}
  \begin{column}{0.25\textwidth}
  \end{column}
\end{columns}


## Session 1 objectives

\vspace{5.6em}

Main goal: Understand the concepts of SCR surveys and be ready for the practicals

## Concepts

\vspace{5.6em}

1.  Ecological surveys and sampling
2.  Capture-recapture surveys
3.  Spatial capture-recapture surveys

## Population and sample

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/intro1.png}
\end{center}

## Population and sample

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/intro2.png}
\end{center}

## Snow leopards: Individual IDs 

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/snowleopards.png}
\end{center}

## (Non-spatial) capture-recapture (CR)

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/intro3.png}
\end{center}

## (Non-spatial) capture-recapture (CR)

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/intro4.png}
\end{center}

## (Non-spatial) capture-recapture (CR)

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/intro5.png}
\end{center}

## (Non-spatial) capture-recapture (CR)

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/intro6.png}
\end{center}

## (Non-spatial) capture-recapture (CR)

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/intro7.png}
\end{center}

## (Non-spatial) capture-recapture (CR): Capture histories

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/caphist1.png}
\end{center}

## (Non-spatial) capture-recapture (CR): Capture histories

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/caphist2.png}
\end{center}

## Spatial capture-recapture (SCR): Single/multi catch traps

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/scr1.png}
\end{center}

## Spatial capture-recapture (SCR): Single/multi catch traps

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/scr2.png}
\end{center}

## Spatial capture-recapture (SCR): Single/multi catch traps

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/scr3.png}
\end{center}

## Spatial capture-recapture (SCR): Single/multi catch traps

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/scr4.png}
\end{center}

## Spatial capture-recapture (SCR): Count traps

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/scr5.png}
\end{center}

## Spatial capture-recapture (SCR): Activity centre

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/scr6.png}
\end{center}

## Spatial capture-recapture (SCR): Activity centre

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/scr7.png}
\end{center}

## Spatial capture-recapture (SCR): Detector types


: **SCR (`secr`) detector types**

| **Category** | **Type (conceptual)** | **`secr` name** |
|:----|:-----------------------------|:-------|
| **Point** | Single-capture trap (1 animal at a time) | single |
| | Multi-capture trap (multiple animals) | multi |
| | Proximity detector (non-exclusive detection) | proximity |
| | Count detector (multiple detections per occasion) | count |
| | Binary proximity with cap of 1 detection | capped |
| | Acoustic / signal-based detector | signal |
| **Area** | Area search detector | polygon |
| | Area search with independent subdetectors | polygonX |
| **Line** | Linear search detector | transect |
|  | Linear search with independent subdetectors | transectX |


## Spatial capture-recapture (SCR): What we know / observe

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/scr8.png}
\end{center}

## SCR: What we want to know: abundance / density 

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/scr9.png}
\end{center}

## SCR: How `secr` does it: Mask and buffer

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/mask.png}
\end{center}


## SCR: How `secr` does it: Activity centres

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/intensity.png}
\end{center}


## SCR: Models

\begin{center}
\includegraphics[height=2.45in]{code-and-figures/models.png}
\end{center}
\vspace{.6em}


## Spatial capture-recapture: Data
\footnotesize

```{r}
#| eval: true
#| echo: true

head(traps_df, 3)
head(caps_df, 3)
```
\normalsize


## Spatial capture-recapture (SCR): Assumptions

: **SCR Assumptions**

| # | Assumption | Description |
|:--|:----------|:---------------------------|
| 1 | Population closure | No births, deaths, immigration, or emigration occur during the sampling period. |
| 2 | Fixed activity centres | Each individual has a stable activity centre that does not change during the study. |
| 3 | Distance-dependent detection | Detection probability declines monotonically with distance from an individual’s activity centre. |
| 4 | Independence of detections | Detection events are independent across individuals and sampling occasions (conditional on model parameters). |

## Spatial capture-recapture (SCR): Assumptions

: **SCR Assumptions**

| # | Assumption | Description |
|:--|:----------|:---------------------------|
| 5 | Correct individual identification | All individuals are uniquely and correctly identified with no mis-classification errors. |
| 6 | Known detector locations | Trap or detector locations are fixed and known without error. |
| 7 | Detection homogeneity or modeled variation | Detection parameters are either constant across individuals or heterogeneity is explicitly modeled (e.g., via covariates). |
| 8 | Activity centre distribution | Activity centres arise from a defined spatial point process (often uniform or Poisson) within the state space. |
| 9 | Known sampling effort | Sampling effort, including detector operation and timing, is accurately recorded and incorporated into the model. |

## Spatial capture-recapture (SCR): Terminology

\footnotesize

: **SCR terminology and definitions (1)**

| Term | Definition |
|:-------|:----------------------------------------|
| Capture history | A record of detections for each individual across detectors and sampling occasions. |
| Detector (trap) | A device or method used to detect individuals (e.g., camera trap, hair snare, acoustic recorder). |
| Detector array | The spatial configuration of all detectors used in the study. |
| Activity centre | The latent (unobserved) central location of an individual’s home range, assumed fixed during the study. |
| State space | The spatial region that contains all possible activity centres of individuals in the population. |
| Mask | A discretized representation of the state space used in computation (grid of possible activity centre locations). |
| Buffer | An added margin around the detector array to ensure the state space includes all individuals that could be detected. |

\normalsize

## Spatial capture-recapture (SCR): Terminology

\footnotesize

: **SCR terminology and definitions (2)**


| Term | Definition |
|:-------|:----------------------------------------|
| Session | A distinct sampling period with its own detector array, effort, and capture data (e.g., a study site or year). |
| Occasion | A within-session sampling period (e.g., day, night, week) during which detections are recorded. |
| Effort | Information on when and where detectors were active; used to correctly model detection probability. |
| Detection function | A function describing how detection probability declines with distance from the activity centre. |
| g0 | The probability of detecting an individual at its activity centre (i.e., distance = 0). |
| $σ$ (sigma) | Spatial scale parameter governing how quickly detection probability declines with distance. |
| Density (D) | The number of individuals per unit area. |

\normalsize

## Spatial capture-recapture (SCR): Terminology

\footnotesize

: **SCR terminology and definitions (3)**

| Term | Definition |
|:-------|:----------------------------------------|
| Encounter model | The statistical model describing how detections arise from the spatial process and detection process. |
| Spatial point process | The probabilistic model describing the distribution of activity centres across space. |
| Likelihood | The probability of observing the data given model parameters; used for parameter estimation. |
| Inhomogeneous density | A model where density varies across space as a function of environmental covariates. |
| Homogeneous density | A model assuming constant density across the state space. |

\normalsize

