---
title: Adding complexity
subtitle: Namkha advanced tutorial
author: |
  | Snow leopard SECR workflow workshop
  | Ian Durbach and Cornelia Oedekoven
institute: |
  | Centre for Research into Ecological & Environmental Modelling
  | University of St Andrews
fontsize: 10pt
format:
  beamer:
    theme: metropolis
    incremental: false
    slide-level: 2
    pdf-engine: xelatex
    syntax-highlighting: tango
classoption: t
urlcolor: teal
execute:
  echo: true
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(knitr)
library(secr)
library(dplyr)
library(sf)
library(ggplot2)
library(MASS)

source("../../namkha_basic/code/more-utilities.R")

knitr::opts_knit$set(global.par = TRUE)
knitr::opts_chunk$set(fig.path = "code-and-figures/")

def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(
  chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    chunk_size <- if (is.null(options$size)) "normalsize" else options$size
    ifelse(
      chunk_size != "normalsize",
      paste0("\n \\", chunk_size, "\n\n", x, "\n\n \\normalsize"),
      x
    )
  }
)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines) == 1) {
    if (length(x) > lines) {
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})

load("../output/namkha_advanced_secr_inputs_capthist.RData")
load("../output/namkha_advanced_secr_inputs_mask.RData")
load("../output/namkha_advanced_secr_inputs_msmask.RData")
load("../output/namkha_advanced_fitted_models_full.RData")
load("../output/namkha_advanced_fitted_models_short.RData")
load("../output/namkha_advanced_fitted_models_ms.RData")

sex_env <- new.env()
load("../output/namkha_advanced_fitted_models_full_sex.RData", envir = sex_env)
sex_f0 <- sex_env$f0
sex_f1 <- sex_env$f1
sex_f2 <- sex_env$f2
sex_f3 <- sex_env$f3
sex_f4 <- sex_env$f4
sex_f0g <- sex_env$f0g
sex_f1g <- sex_env$f1g
sex_f2g <- sex_env$f2g
sex_f3g <- sex_env$f3g
sex_f4g <- sex_env$f4g

full_results <- read.csv("../output/results/full/namkha_advanced_results_abundance_AIC_full.csv")
short_results <- read.csv("../output/results/short/namkha_advanced_results_abundance_AIC_short.csv")
ms_results <- read.csv("../output/results/ms/namkha_advanced_results_abundance_AIC_ms.csv")
```

## Session 3 objectives

Main goal: Show where the simple workflow has to change when the analysis gets more realistic.

1. Longer survey periods and closure
2. Shortened-survey and multisession alternatives
3. Elevation constraints and reporting masks
4. Gorge-avoiding movement distances
5. Sex-specific model components

\vspace{0.4cm}

\alert{Assumption:} You have already seen the simple workflow from Session 2.

## From simple to advanced

The basic workflow stays the same:

\begin{tikzpicture}[remember picture, overlay]
\node[xshift=-4.5cm, yshift=-0.8cm, draw=black, fill=blue!10,
      fill opacity = 0.3, text opacity = 1, shape=rectangle,
      rounded corners=1.5] at (current page){
      \begin{minipage}[t][3.5cm][t]{0.15\textwidth}
       \textbf{01}\\
       camera file\\
       detection file\\
       $\Rightarrow$ \\
       \texttt{capthist}
      \end{minipage}};
\node[xshift=-2.25cm, yshift=-0.8cm, draw=black, fill=blue!10,
      fill opacity = 0.3, text opacity = 1, shape=rectangle,
      rounded corners=1.5] at (current page){
      \begin{minipage}[t][3.5cm][t]{0.15\textwidth}
       \textbf{02}\\
       survey area\\
       buffer\\
       covariates\\
       $\Rightarrow$ \\
       \texttt{mask}
      \end{minipage}};
\node[xshift=0cm, yshift=-0.8cm, draw=black, fill=blue!10,
      fill opacity = 0.3, text opacity = 1, shape=rectangle,
      rounded corners=1.5] at (current page){
      \begin{minipage}[t][3.5cm][t]{0.17\textwidth}
       \textbf{03}\\
       (crop for \\ 
       multisession \\
       model) \\
       $\Rightarrow$ \\
       \texttt{mask (ms)}
      \end{minipage}};
\node[xshift=2.25cm, yshift=-0.8cm, draw=black, fill=blue!10,
      fill opacity = 0.3, text opacity = 1, shape=rectangle,
      rounded corners=1.5] at (current page){
      \begin{minipage}[t][3.5cm][t]{0.15\textwidth}
       \textbf{04}\\
       $D$ model\\
       $\lambda_0$/$g_0$ model\\
       $\Rightarrow$ \\
       \texttt{secr.fit}
      \end{minipage}};
\node[xshift=4.5cm, yshift=-0.8cm, draw=black, fill=blue!10,
      fill opacity = 0.3, text opacity = 1, shape=rectangle,
      rounded corners=1.5] at (current page){
      \begin{minipage}[t][3.5cm][t]{0.15\textwidth}
       \textbf{05}\\
       AICc\\
       abundance\\
       density\\
       maps
      \end{minipage}};
\end{tikzpicture}

Each extension affects a specific part of the workflow.

# Extension 1: Long surveys and closure

## Extension 1: Long surveys and closure

\alert{Problem:} the full survey is long enough that the closed-population assumption may be questionable.

\alert{Implementation:}

- keep the full survey
- create a shortened survey window
- split the full survey into temporal sessions
- compare how results change

\vspace{0.3cm}

Scripts involved:

- `01_make_capthist.R`
- `03_make_ms_masks.R`
- `04a`, `04b`, `04c`
- `05a`, `05b`, `05c`

## Full survey: occasion-specific usage

`secr` implements closure checks. These need multiple occasions (no other need for these i.e. for model fitting)

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
full_occ_length <- 14 
full_occ_start <- seq.Date(from = survey_start, to = survey_end,
                           by = paste(full_occ_length, "days"))
full_occ_bins <- interval(start = full_occ_start, 
                          end = full_occ_end)
```
\normalsize

For each camera and occasion, need to recalculate effort (days active)

\vspace{.6em}
\footnotesize
```{r}
#| code-line-numbers: "1|4"
#| eval: false
#| echo: true
usage_full <- matrix(0, nrow = nrow(cameras), ncol = length(full_occ_bins))
for (i in seq_len(nrow(cameras))) {
  usage_full[i, ] <- round(as.numeric(
    int_overlaps_numeric(cameras$camera_on[i], full_occ_bins)) / (24 * 60 * 60))
}
usage(full_traps) <- usage_full
```
\normalsize

## Full survey: descriptive check

\begin{columns}[T]
  \begin{column}{0.7\textwidth}
    \centering
    \includegraphics[width=\linewidth]{../output/fig/namkha_capture_history_summary.png}
  \end{column}

  \begin{column}{0.3\textwidth}
  \vspace{1.7em}
    \begin{itemize}
      \item cameras started at different times
      \item new animals appear through time
      \item shortened window is a defensible alternative
      \item session split is another alternative
    \end{itemize}
  \end{column}
\end{columns}

## Full survey: closure checks

The full survey occasions allow a formal closure check

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
closure.test(full_ch)
```
\normalsize

\vspace{.6em}

```{r}
#| eval: true
#| echo: false
#| output: true
closure.test(full_ch)
```
\normalsize

Suggests moderate concern about closure. Now check closure for the shortened survey

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
short_occasion_start <- floor(as.numeric(short_survey_start - survey_start) /
                              full_occ_length) + 1
short_occasion_end <- floor(as.numeric(short_survey_end - survey_start) /
                            full_occ_length) + 1

closure.test(subset(full_ch, occasions = short_occasion_start:short_occasion_end))
```
\normalsize

```{r}
#| eval: true
#| echo: false
#| output: true
short_occasion_start <- floor(as.numeric(short_survey_start - survey_start) /
                              full_occ_length) + 1
short_occasion_end <- floor(as.numeric(short_survey_end - survey_start) /
                            full_occ_length) + 1

closure.test(subset(full_ch, occasions = short_occasion_start:short_occasion_end))
```

Much less (no) evidence against closure.

## Shortened survey: workflow change

\alert{Key idea:} limit survey length by taking a subset of the full survey 

\alert{Issue:} Which part of the full survey to take?

\vspace{.6em}

Keep only detections in the shortened window

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
short_capts_raw <- capts_raw |>
  filter(date >= short_survey_start, date <= short_survey_end)
```
\normalsize

Recalculate camera effort for that same window

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
short_traps_df <- cameras |>
  mutate(
    effort_short = pmax(
      0,
      as.numeric(pmin(end_date, short_survey_end) -
                 pmax(start_date, short_survey_start), units = "days")
    )
  )
```
\normalsize

## Shortened survey: `capthist`

Create and check the capture history

\vspace{.6em}
\footnotesize
```{r}
#| code-line-numbers: "1|2|3"
#| eval: false
#| echo: true
short_capts <- short_capts_raw |>
  mutate(occasion = 1L) |>
  dplyr::select(session, animalID, occasion, trapID, sex)

short_ch <- make.capthist(captures = short_capts, traps = short_traps)
verify(short_ch)
summary(short_ch, terse = TRUE)
```
\normalsize

\vspace{.6em}

```{r}
#| eval: true
#| echo: false
#| output: true
summary(short_ch, terse = TRUE)
```

## Multisession survey: workflow change

\alert{Key idea:} keep the full survey but split it into temporal sessions.

Detections get a session label:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
ms_capts <- capts_raw |>
  mutate(
    session = if_else(date < ms_session_end, 1L, 2L),
    occasion = 1L
  ) |> dplyr::select(session, animalID, occasion, trapID, sex)
```
\normalsize

Camera effort is calculated separately for each session:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
ms_traps_df <- cameras |>
  mutate(
    effort_s1 = pmax(0, as.numeric(pmin(end_date, ms_session_end - days(1)) -
                                   pmax(start_date, survey_start), 
                                   units = "days")),
    effort_s2 = pmax(0, as.numeric(pmin(end_date, survey_end) -
                                   pmax(start_date, ms_session_end), 
                                   units = "days"))
  )
```
\normalsize

## Multisession survey: `capthist`

For multisession `secr`, the traps object is a named list:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
ms_traps_list <- split(ms_traps_df, f = ms_traps_df$session)
traps_ms <- vector("list", length(ms_traps_list))

for (i in seq_along(ms_traps_list)) {
  traps_ms[[i]] <- read.traps(
    data = ms_traps_list[[i]] |> dplyr::select(trapID, x, y, effort),
    detector = "count",
    trapID = "trapID",
    binary.usage = FALSE
  )
}
names(traps_ms) <- c("1", "2")
```
\normalsize

Then add covariates, etc.

## Multisession survey: factor levels

Create and check the multisession capture history:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
ch_ms <- make.capthist(captures = ms_capts, traps = traps_ms)
verify(ch_ms)
```
\normalsize

Make sure to share factor levels across sessions:

\vspace{.6em}
\footnotesize
```{r}
#| code-line-numbers: "1|2"
#| eval: false
#| echo: true
ch_ms <- shareFactorLevels(ch_ms, stringsAsFactors = FALSE)
verify(ch_ms)
summary(ch_ms, terse = TRUE)
```
\normalsize

\vspace{.6em}

```{r}
#| eval: true
#| echo: false
#| output: true
t(summary(ch_ms, terse = TRUE))
```
\normalsize

# Extension 2: Elevation constraints and reporting masks

## Extension 2: Elevation constraints

\alert{Problem:} not every point in the trap buffer is plausible snow leopard habitat.

\alert{Implementation:}

- add elevation to the full mask
- restrict mask points to plausible elevation range
- keep separate masks for fitting and reporting
- save scaling information for covariate effect plots

## Add elevation to the mask

Elevation is downloaded and saved if not available already:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
elev_file <- "namkha_advanced/output/cache/namkha_elevation.tif"

if (file.exists(elev_file)) {
  elev <- rast(elev_file)
} else {
  elev_download <- get_elev_raster(locations = full_region_4326,
                                   z = 10, clip = "locations")
  elev <- rast(elev_download)
  writeRaster(elev, elev_file, overwrite = TRUE)
}
```
\normalsize

Then project and attach it to the mask:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
elev <- project(elev, y = my_crs)
names(elev) <- "elev"
mask <- addCovariates(mask, elev)
```
\normalsize

## Restrict masks by elevation

We use 3000--5500 m as the plausible elevation range.

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
inside_elev_range <- covariates(mask)$elev >= elev_min &
  covariates(mask)$elev <= elev_max
mask_elev <- subset(mask, subset = inside_elev_range)
```
\normalsize

Then create reporting and fitting masks.

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
mask_survey_area_elev <- subset(mask_elev, subset = inside_survey_area_elev)
mask_model <- subset(mask_elev, subset = inside_buffer_elev)
```
\normalsize

## Fitting mask versus reporting mask

\alert{Fitting mask:} `mask_model`

- elevation-constrained
- within the trap buffer
- can include activity centres outside Namkha

\vspace{0.3cm}

\alert{Reporting mask:} `mask_survey_area_elev`

- elevation-constrained
- inside the Namkha survey boundary
- used for abundance, density, maps, and covariate summaries

\vspace{0.3cm}

Same conceptual split as Session 2, but now with an extra habitat constraint.

# Extension 3: Barriers and custom movement distances

## Extension 3: Gorge-avoiding movement

\alert{Problem:} Euclidean distance says animals can move straight through the gorge.

\alert{Implementation:}

- read the gorge geometry
- make a fine mask for path calculations
- calculate shortest-path distances with `nedist()`
- pass those distances to `secr.fit()` using `details`

## Read the gorge

The gorge is a polygon that must be in the same CRS as traps and masks.

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
gorge <- st_read(".../namkha_gorge.kml") |>
  st_transform(my_crs) |>
  st_geometry()
```
\normalsize

\begin{center}
\includegraphics[height=2in]{../output/results/full/namkha_advanced_survey_region_full.png}
\end{center}

## Fine mask for movement distances

Gorge is narrower than existing mask spacing, so shortest-path calculations will be inaccurate. Use a finer mask to calculate distances around the gorge.

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
mask_fine <- make.mask(
  traps = full_traps,
  buffer = mask_buffer,
  type = "trapbuffer",
  spacing = 200,
  poly = gorge,
  poly.habitat = FALSE,
  keep.poly = FALSE
)
```
\normalsize

\alert{Note:} Still use the normal model mask for model fitting.

## Non-Euclidean distance matrix

`nedist()` calculates distances from traps to mask points, using the fine mask as the movement network.

\vspace{.6em}
\footnotesize
```{r}
#| code-line-numbers: "1|5"
#| eval: false
#| echo: true
userd <- nedist(
  full_traps,
  mask,
  mask_fine,
  directions = 16
)
```
\normalsize

Each cropped mask needs its own `userdist` (this is just a subsetting of rows and columns of the full `userdist`)

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
userd_elev <- userd[, inside_elev_range, drop = FALSE]
userd_model <- userd_elev[, inside_buffer_elev, drop = FALSE]
userd_survey_area_elev <- userd_elev[, inside_survey_area_elev, drop = FALSE]
```
\normalsize

## Non-Euclidean distance check

\begin{columns}[T] % align at top
  \begin{column}{0.7\textwidth}
    \centering
    \includegraphics[width=\linewidth]{code-and-figures/plot-userdist-example-1.png}
  \end{column}

  \begin{column}{0.3\textwidth}
  \vspace{2em}
    \begin{itemize}
      \item distance from camera `TPC13` to each mask point
      \item points across the gorge are farther away than straight-line distance suggests
    \end{itemize}
  \end{column}
\end{columns}

## Use `userdist` in model fitting

This is the key change to `secr.fit()`:

\vspace{.6em}
\footnotesize
```{r}
#| code-line-numbers: "6"
#| eval: false
#| echo: true
f0 <- secr.fit(
  full_ch,
  detectfn = "HHN",
  mask = mask_model,
  model = list(D ~ 1, lambda0 ~ 1, sigma ~ 1),
  details = list(userdist = userd_model),
  verify = FALSE,
  trace = FALSE
)
```
\normalsize

The model still uses `mask_model`; only the movement distances have changed.

# Extension 4: Multisession masks and distances

## Extension 4: Multisession masks

\alert{Problem:} `ch_ms` has multiple sessions, so model fitting needs one mask per session.

\alert{Implementation:}

- build a trap-buffer polygon for each session
- crop the elevation-constrained mask to each session
- create one `userdist` matrix per session
- keep names aligned with `traps_ms`

## One mask per session

Create polygons around traps active in each session:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
mask_per_session_sf <- list()
for (i in seq_along(traps_ms)) {
  mask_per_session_sf[[i]] <- st_as_sf(traps_ms[[i]],
                                       coords = c("x", "y"), crs = my_crs) |>
    st_buffer(mask_buffer) |>
    st_union()
}
names(mask_per_session_sf) <- names(traps_ms)
```
\normalsize

Crop the full elevation mask:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
mask_ms <- list()
for (i in seq_along(mask_per_session_sf)) {
  mask_in <- lengths(st_intersects(mask_elev_sf,
                                   mask_per_session_sf[[i]])) > 0
  mask_ms[[i]] <- subset(mask_elev, subset = mask_in)
}
names(mask_ms) <- names(traps_ms)
```
\normalsize

## One `userdist` per session

Use the same mask indicator to subset the full distance matrix:

\vspace{.6em}
\footnotesize
```{r}
#| code-line-numbers: "1|5"
#| eval: false
#| echo: true
userd_ms <- list()
for (i in seq_along(mask_per_session_sf)) {
  mask_in <- lengths(st_intersects(mask_elev_sf,
                                   mask_per_session_sf[[i]])) > 0
  userd_ms[[i]] <- userd_elev[, mask_in, drop = FALSE]
}
names(userd_ms) <- names(traps_ms)
```
\normalsize

The names of `mask_ms`, `userd_ms`, and `traps_ms` should all match.

## Fit multisession models

The fitting call uses lists for `mask` and `userdist`:

\vspace{.6em}
\footnotesize
```{r}
#| code-line-numbers: "4|6"
#| eval: false
#| echo: true
ms0 <- secr.fit(
  ch_ms,
  detectfn = "HHN",
  mask = mask_ms,
  model = list(D ~ 1, lambda0 ~ 1, sigma ~ 1),
  details = list(userdist = userd_ms),
  verify = FALSE,
  trace = FALSE
)
```
\normalsize

Multisession analyses also good when sessions are spatial (geographically separate surveys that you want to combine)

## Multisession model set

Can share parameters across sessions:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
ms1 <- secr.fit(ch_ms, detectfn = "HHN", mask = mask_ms,
                model = list(D ~ std_tri, lambda0 ~ 1, sigma ~ 1),
                details = list(userdist = userd_ms), start = ms0)
```
\normalsize

Or include session as a covariate:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
ms5 <- secr.fit(ch_ms, detectfn = "HHN", mask = mask_ms,
                model = list(D ~ session, lambda0 ~ 1, sigma ~ 1),
                details = list(userdist = userd_ms), start = ms0)
```
\normalsize

\alert{Note:} How to extrapolate from session-specific covariate models to a wider region is unclear.

# Extension 5: Sex-specific models

## Extension 5: Sex-specific models

\alert{Problem 1:} detection or density may differ by sex 
\alert{Problem 2:} sex not known for all animals

Two approaches:

1. `hcov = "sex"`: individual covariate models (Problem 1 & 2)
2. `groups = "sex"`: grouped real-parameter models (Problem 1 only)

\vspace{0.3cm}

\alert{Important:} do not compare the `hcov` and `groups` models with each other (or with other secr models) using AIC.

## Sex as an individual covariate

`hcov = "sex"` are finite mixture models that retain animals with unknown sex.

\vspace{.6em}
\footnotesize
```{r}
#| code-line-numbers: "5|6"
#| eval: false
#| echo: true
f0 <- secr.fit(
  full_ch,
  detectfn = "HHN",
  mask = mask_model,
  model = list(D ~ 1, lambda0 ~ h2, sigma ~ 1),
  hcov = "sex",
  details = list(userdist = userd_model),
  verify = FALSE
)
```
\normalsize

`h2` is the term used by `secr` for the first two levels of the individual covariate.

## Sex as groups

Grouped models estimate separate real parameters by sex, but animals with unknown sex must be removed:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
full_ch_known_sex <- subset(
  full_ch,
  subset = covariates(full_ch)$sex != "Unknown"
)
full_ch_known_sex <- shareFactorLevels(full_ch_known_sex)
```
\normalsize

Then fit grouped models:

\vspace{.6em}
\footnotesize
```{r}
#| code-line-numbers: "5"
#| eval: false
#| echo: true
f0g <- secr.fit(
  full_ch_known_sex,
  detectfn = "HHN",
  mask = mask_model,
  model = list(D ~ 1, lambda0 ~ g, sigma ~ 1),
  groups = "sex",
  details = list(userdist = userd_model)
)
```
\normalsize

## Sex-specific model comparison

Compare each sex-specific candidate set internally:

\vspace{.6em}
\footnotesize
```{r}
#| eval: true
#| echo: true
AIC(sex_f0, sex_f1, sex_f2, sex_f3, sex_f4, criterion = "AICc")
```
\normalsize

## Sex-specific model comparison

Compare each sex-specific candidate set internally:

\vspace{.6em}
\footnotesize
```{r}
#| eval: true
#| echo: true
AIC(sex_f0g, sex_f1g, sex_f2g, sex_f3g, sex_f4g, criterion = "AICc")
```
\normalsize

## Sex-specific model comparison

Finite mixtures, best model:

\footnotesize
```{r}
#| eval: true
#| echo: true
predict(sex_f3)
```

## Sex-specific model comparison

Groups, best model:

\footnotesize
```{r}
#| eval: true
#| echo: true
predict(sex_f3g)
```

# Extension 6: Advanced reporting

## Extension 6: Advanced reporting

Reporting follows the Session 2 pattern:

- choose fitted models
- choose reporting mask
- estimate abundance with `region.N()`
- convert abundance to density
- join with AICc
- make maps and covariate-effect plots

\vspace{0.3cm}

\alert{Main change:} If reporting over a new mask, replace `userdist` with the matching distance matrix.

Other reporting code works more or less like before (see scripts and vignette for details).

## Reporting mask and `userdist`

Choose the reporting mask:

\vspace{.6em}
\footnotesize
```{r}
#| eval: false
#| echo: true
extrap_mask <- mask_survey_area_elev
extrap_mask_userd <- userd_survey_area_elev
```
\normalsize

Copy the fitted models and swap the distance matrix:

\vspace{.6em}
\footnotesize
```{r}
#| code-line-numbers: "2"
#| eval: false
#| echo: true
fitted_models_region <- lapply(fitted_models, function(model) {
  model$details$userdist <- extrap_mask_userd
  model
})
```
\normalsize

Now fitted models stored in `fitted_models_region` can be used to extrapolate to the reporting mask (e.g. abundance, maps).

## Comparing full, short, multisession surveys

Full survey:

\scriptsize
```{r}
#| eval: true
#| echo: false
#| output: true
full_results |>
  dplyr::select(modelname, AICc, dAICc, estimate, D, CV) |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))
```
\normalsize

Short survey:

\vspace{.4em}
\scriptsize
```{r}
#| eval: true
#| echo: false
#| output: true
short_results |>
  dplyr::select(modelname, AICc, dAICc, estimate, D, CV) |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))
```
\normalsize

Multisession survey may return session-specific results (not here).

\scriptsize
```{r}
#| eval: true
#| echo: false
#| output: true
ms_results |>
  filter(session == 1) |> 
  dplyr::select(modelname, AICc, dAICc, estimate, D, CV) |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))
```
\normalsize

## Predicted density maps

\begin{columns}[T]
  \begin{column}{0.5\textwidth}
    \centering
    Full survey

    \includegraphics[width=\linewidth]{../output/results/full/namkha_advanced_predicted_density_full.png}
  \end{column}
  \begin{column}{0.5\textwidth}
    \centering
    Shortened survey

    \includegraphics[width=\linewidth]{../output/results/short/namkha_advanced_predicted_density_short.png}
  \end{column}
\end{columns}

## Predicted density maps

\begin{columns}[T]
  \begin{column}{0.5\textwidth}
    \centering
    Multisession session 1

    \includegraphics[width=\linewidth]{../output/results/ms/namkha_advanced_predicted_density_ms_session_1.png}
  \end{column}
  \begin{column}{0.5\textwidth}
    \centering
    Multisession session 2

    \includegraphics[width=\linewidth]{../output/results/ms/namkha_advanced_predicted_density_ms_session_2.png}
  \end{column}
\end{columns}

## Workflow recap

\vspace{1em}

1. Long surveys require thinking about closure. \vspace{0.7em}
2. Shortened surveys and multisession analyses change both captures and effort. \vspace{0.7em}
3. Habitat constraints change masks for fitting and reporting. \vspace{0.7em}
4. Barriers change movement distances, not the capture-history object. \vspace{0.7em}
5. Sex-specific models require careful handling of unknown sex and AIC comparisons. \vspace{0.7em}
6. Extrapolation to a new mask needs a `userdist` that matches.
